---
title: "TP1 — Exploring Biomedical Data"
subtitle: "Doctors as Data Scientists · Introduction to Exploratory Data Analysis in R"
author: "Santiago with Claude"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, fig.align = "center")
```

> **How to use this notebook.** You do **not** need any programming experience.
> Read each short explanation, then press the green *play* button (▶) at the
> top-right of each grey *code chunk* to run it. Look at the output, and answer
> the **Questions** in the green boxes (write your answers in plain text under
> each one). The goal is to *understand the data*, not to memorise code.
>
> This practical has **two case studies**. We warm up on intuitive body
> measurements (Case Study 1), then apply the same ideas — and go deeper — on a
> real clinical dataset (Case Study 2).

## Set-up (run once)

```{r install, eval=FALSE}
# Run this ONCE. If a package is already installed nothing bad happens.
install.packages(c("corrplot",        # correlation pictures
                   "scatterplot3d"))  # 3-D scatter plot
```

```{r libraries}
library(corrplot)
library(scatterplot3d)
```

---

# Case Study 1 — Body dimensions of healthy adults

We start with an easy, intuitive dataset: **body measurements of 507 physically
active adults** (247 men and 260 women), from the
[openintro](https://www.openintro.org/data/index.php?data=bdims) package. Every
column is something you can picture with a tape measure.

The 21 body measurements come in two families (all in **cm**):

| Family | Examples (column codes) | Meaning |
|--------|-------------------------|---------|
| **Skeletal diameters** | `bia_di`, `bii_di`, `bit_di`, `che_di`, `elb_di`, `wri_di`, `kne_di`, `ank_di` | bone-to-bone breadths (shoulder, pelvis, chest, elbow, wrist, knee, ankle) |
| **Girths (circumferences)** | `sho_gi`, `che_gi`, `wai_gi`, `hip_gi`, `thi_gi`, `bic_gi`, `for_gi`, `cal_gi` | shoulder, chest, waist, hip, thigh, biceps, forearm, calf |

Plus `age` (years), `wgt` (weight, kg), `hgt` (height, cm) and `sex`
(1 = male, 0 = female).

## Load the data

```{r load-bdims}
bd <- read.csv("../data/bdims.csv")
bd$sex <- factor(bd$sex, levels = c(0, 1), labels = c("female", "male"))
head(bd)
```

## Getting to know the data

```{r bdims-structure}
dim(bd)          # rows (people) and columns (measurements)
table(bd$sex)    # how many women and men
```

> **Question 1.1.** How many people are in the dataset, and how many measurements
> were taken on each one? Which single variable is *categorical* (a label) rather
> than a number? What are the units of `hgt`, `wgt` and the girths?

## What shape is a measurement?

```{r bdims-hist, fig.width=9}
par(mfrow = c(1, 2))
hist(bd$hgt, breaks = 30, col = "steelblue",  main = "Height", xlab = "height (cm)")
hist(bd$wgt, breaks = 30, col = "darkorange", main = "Weight", xlab = "weight (kg)")
par(mfrow = c(1, 1))
```

> **Question 1.2.** Are height and weight roughly *symmetric* (bell-shaped), or
> strongly lop-sided? (Keep this in mind: in Case Study 2 the clinical lab values
> will look very different.)

> **Question 1.3.** Make a boxplot of weight (`boxplot(bd$wgt)`). Are there a few
> unusually heavy or light people? In a sample of healthy adults, would you expect
> many extreme outliers?

## Comparing the two groups (men vs women)

```{r bdims-sex, fig.width=9}
par(mfrow = c(1, 2))
boxplot(sho_gi ~ sex, data = bd, col = c("plum","steelblue"),
        main = "Shoulder girth", xlab = "", ylab = "cm")
boxplot(hip_gi ~ sex, data = bd, col = c("plum","steelblue"),
        main = "Hip girth", xlab = "", ylab = "cm")
par(mfrow = c(1, 1))
```

> **Question 1.4.** Which measurement separates men and women more clearly —
> shoulder girth or hip girth? Try a few others (e.g. `wri_di`, `thi_gi`). Which
> body measurements are most "male/female different", and which barely differ?

## Relationships between measurements

```{r bdims-cor}
meas <- c("hgt","wgt","sho_gi","che_gi","wai_gi","hip_gi","thi_gi","bic_gi")
round(cor(bd[, meas]), 2)
```

```{r bdims-scatter}
plot(bd$wgt, bd$che_gi, pch = 19, col = rgb(0,0,1,0.3),
     main = "Chest girth vs weight", xlab = "weight (kg)", ylab = "chest girth (cm)")
```

> **Question 1.5.** Which pair of measurements is most strongly correlated? Are any
> pairs *negatively* correlated — and does it make sense that, in body
> measurements, almost everything grows together?

## Putting measurements on a common scale

```{r bdims-scale}
bd_z <- scale(bd[, meas])               # mean 0, sd 1 for each measurement
round(colMeans(bd_z), 2)                # all ~0
round(apply(bd_z, 2, sd), 2)            # all 1
```

> **Question 1.6.** Weight is in kilograms, the girths in centimetres. Why does
> standardising (mean 0, sd 1) let us compare and combine them fairly? (We rely on
> this in TP2 and TP3.)

---

# Case Study 2 — Hepatitis C blood panel

Now a real clinical dataset. Each row is one person who had a blood sample
analysed. The data come from the
[UCI HCV dataset](https://archive.ics.uci.edu/dataset/571/hcv+data): healthy
**blood donors** and patients with hepatitis C at increasing stages of liver
damage (**Hepatitis → Fibrosis → Cirrhosis**).

The numeric columns are routine **laboratory tests**:

| Code | Lab test | What it reflects |
|------|----------|------------------|
| ALB  | Albumin | A protein the liver makes (synthetic function) |
| ALP  | Alkaline phosphatase | Bile ducts / liver |
| ALT  | Alanine aminotransferase | Liver-cell damage |
| AST  | Aspartate aminotransferase | Liver-cell damage |
| BIL  | Bilirubin | Liver clearance (rises in jaundice) |
| CHE  | Cholinesterase | Liver synthetic function |
| CHOL | Cholesterol | Liver / metabolism |
| CREA | Creatinine | Kidney function |
| GGT  | Gamma-glutamyl transferase | Bile ducts / liver |
| PROT | Total protein | Overall protein level |

`Category` is the diagnosis and `Sex` is m/f.

> **A note on units.** The source dataset does not state the measurement units.
> Based on the value ranges and the original publication, they are assumed to be
> the conventional clinical units — enzymes (ALT, AST, ALP, GGT) in U/L,
> bilirubin and creatinine in µmol/L, albumin and total protein in g/L,
> cholinesterase in kU/L, and cholesterol in mmol/L. We show "U/L" on some plot
> axes on that assumption; treat it as inferred, not documented.

## Load the data

```{r load}
hcv <- read.csv("../data/hcvdat0.csv", row.names = 1)
head(hcv)
```

We also build a short, readable diagnosis label we will reuse throughout, and a
vector with the names of the 10 lab tests.

```{r setup-helpers}
hcv$Group <- factor(gsub("^0.*", "Donor", hcv$Category),
                    levels = c("Donor","1=Hepatitis","2=Fibrosis","3=Cirrhosis"),
                    labels = c("Donor","Hepatitis","Fibrosis","Cirrhosis"))

labs <- c("ALB","ALP","ALT","AST","BIL","CHE","CHOL","CREA","GGT","PROT")
```

## Part A — Getting to know the data

```{r structure}
# How many rows (patients) and columns (variables)?
dim(hcv)

# What is the TYPE of each variable?
str(hcv)
```

`int`/`num` columns are **numeric** (the lab values and age). `chr`/`factor`
columns are **categorical** (`Category`, `Sex`, `Group`).

```{r categories}
table(hcv$Category)   # patients per diagnosis
table(hcv$Sex)        # patients per sex
```

> **Question 2.1.** How many patients are in the dataset, and how many variables
> were measured on each one? Which variables are numeric measurements and which
> are categorical labels?

> **Question 2.2.** How many patients are healthy blood donors, and how many have
> some form of liver disease (Hepatitis, Fibrosis, Cirrhosis)? Is the dataset
> *balanced* between healthy and sick people?

### Looking at the raw file before trusting it

Before any analysis, a good data scientist *opens the raw file* to see how it is
written. The chunk below prints the first three lines exactly as stored on disk.

```{r raw-lines}
readLines("../data/hcvdat0.csv", n = 3)
```

> **Question 2.3.** From those raw lines, work out: (a) what character **separates
> the columns** (a comma? a semicolon? a tab?); (b) what character is the
> **decimal point** (a dot or a comma?); and (c) how a **missing value** would
> appear (here, the letters `NA`). Why does R need to know the separator and the
> decimal mark to read the file correctly? (This is why we used `read.csv`, which
> assumes a comma separator and a dot decimal.)

### Missing values

Real medical data almost always has gaps. In R a missing value is shown as `NA`.

```{r missing}
colSums(is.na(hcv))           # how many NA in each column
sum(!complete.cases(hcv))     # how many patients have at least one NA
```

> **Question 2.4.** Which lab tests have missing values, and how many? If you
> wanted to analyse all 10 lab tests together, what could you do about the
> patients with missing results? (Two common options: drop those patients, or
> fill the gaps with an estimate.)

## Part B — Describing one variable at a time

```{r summary}
# A quick five-number summary of every variable
summary(hcv[, labs])
```

```{r mean-median}
mean(hcv$AST, na.rm = TRUE)    # na.rm = TRUE means "ignore missing values"
median(hcv$AST, na.rm = TRUE)
sd(hcv$AST, na.rm = TRUE)      # standard deviation = spread
```

When the **mean is much larger than the median**, a few very high values are
pulling the average up — the distribution is *right-skewed*.

> **Question 2.5.** Using the `summary()` output, list the lab tests for which the
> mean is clearly larger than the median. What does that tell you about the shape
> of those distributions? (Contrast with the body measurements in Case Study 1.)

### Histograms and boxplots

```{r hist-ast}
hist(hcv$AST, breaks = 40, col = "steelblue",
     main = "Distribution of AST", xlab = "AST (U/L)")
```

```{r box-ast}
boxplot(hcv$AST, horizontal = TRUE, col = "steelblue",
        main = "Boxplot of AST", xlab = "AST (U/L)")
```

> **Question 2.6.** Describe the shape of the AST distribution. Is it symmetric
> (bell-shaped), or does it have a long tail on one side? Now make a histogram for
> `CHE` (change `hcv$AST` to `hcv$CHE`). How is its shape different?

### Spotting unusual values (outliers)

```{r outliers}
max(hcv$CREA, na.rm = TRUE)                                  # most extreme creatinine
hcv[which.max(hcv$CREA), c("Group","Age","Sex","CREA")]     # who is it?
```

> **Question 2.7.** Several lab values are far higher than the rest. Are these
> likely to be *mistakes* (typos, machine errors) or *real* extreme results from
> very sick patients? How would you decide? (Hint: look at the patient's
> diagnosis above.)

## Part C — Distribution shape, normality, and transformations

Many statistical methods assume data are roughly **normal** (bell-shaped). A
quick visual check is the **Q-Q plot**: if the points fall along the straight
line the data are close to normal; if they curve away at the ends, they are not.

```{r qqnorm}
qqnorm(hcv$AST, main = "Q-Q plot of AST"); qqline(hcv$AST, col = "red")
```

### Comparing transformations

For right-skewed lab data we can try the **logarithm**, the **square root**, or
the **reciprocal** (1/x). Let's see which makes AST most symmetric.

```{r transforms, fig.width=10, fig.height=3.4}
par(mfrow = c(1, 4))
hist(hcv$AST,        breaks = 40, col = "grey70",    main = "AST (original)", xlab = "")
hist(sqrt(hcv$AST),  breaks = 40, col = "lightgreen",main = "sqrt(AST)",      xlab = "")
hist(log(hcv$AST),   breaks = 40, col = "darkorange",main = "log(AST)",       xlab = "")
hist(1/hcv$AST,      breaks = 40, col = "plum",      main = "1 / AST",        xlab = "")
par(mfrow = c(1, 1))
```

> **Question 2.8.** Which transformation makes AST look most bell-shaped: square
> root, logarithm, or reciprocal? Check your choice by also drawing its Q-Q plot
> (for example `qqnorm(log(hcv$AST)); qqline(log(hcv$AST))`) — the best
> transformation is the one whose points sit closest to the straight line.

> **Question 2.9.** Repeat the comparison for `GGT` and `BIL`. Why might it be
> convenient to analyse these strongly skewed tests on a transformed scale?
> (Note: this is an *empirical* choice — the transformation that makes the data
> most symmetric — not a rule fixed by the measurement units.)

## Part D — Comparing disease groups

```{r box-by-group, fig.width=9}
boxplot(AST ~ Group, data = hcv, col = "lightblue",
        main = "AST by diagnosis", xlab = "", ylab = "AST (U/L)")
```

> **Question 2.10.** Does AST (a marker of liver-cell damage) tend to increase as
> the disease becomes more severe (donor → hepatitis → fibrosis → cirrhosis)?

> **Question 2.11.** Albumin (`ALB`) and cholinesterase (`CHE`) reflect how well
> the liver is *working*. Make boxplots of `ALB` and `CHE` by `Group`. Do they go
> **up** or **down** in cirrhosis? Does that make biological sense?

### All variables at once: a profile of the means

To compare all ten tests across the four groups in one picture, we first need to
put them on a common scale. The tests use very different units and ranges (CREA
in the hundreds, CHOL below 10), so we **standardise** them: subtract each test's
mean and divide by its standard deviation, giving every test mean 0 and sd 1. The
`scale()` function does this. (We will return to *why* standardising matters in
Part G; for now just treat it as "putting every test on the same ruler".)

```{r profile-means, fig.width=8.5, fig.height=5}
# 1) Put all ten tests on the same scale (mean 0, sd 1)
hcv_z <- scale(hcv[, labs])

# 2) Average each test within each diagnosis group, and look at the table
profiles <- aggregate(hcv_z, by = list(Group = hcv$Group), FUN = mean, na.rm = TRUE)
profiles

# 3) Draw one line per group across the ten tests
matplot(t(profiles[, labs]), type = "b", pch = 19, col = 1:4, xaxt = "n",
        xlab = "", ylab = "standardised mean (0 = overall average)",
        main = "Average lab profile of each diagnosis group")
axis(1, at = 1:length(labs), labels = labs, las = 2)  # write the test names below
legend("topright", legend = profiles$Group, col = 1:4, pch = 19)
abline(h = 0, lty = 3)                                 # the overall average
```

> **Question 2.12.** Reading the profile plot, which lab tests pull the **cirrhosis**
> line far away from the **donor** line (the most *discriminating* tests)? Which
> tests are roughly the same in every group (not useful for telling them apart)?

### Every variable on its own scale

The profile above used standardised averages. It is also worth seeing the raw
distributions, one boxplot per test.

```{r all-boxes, fig.width=10, fig.height=8}
par(mfrow = c(3, 4), mar = c(3, 3, 2, 1))
for (v in labs) {
  boxplot(hcv[[v]] ~ hcv$Group, main = v, xlab = "", ylab = "",
          col = "lightblue", cex.axis = 0.7)
}
par(mfrow = c(1, 1))
```

> **Question 2.13.** Scanning all 10 boxplots, which tests separate the disease
> groups most clearly, and which hardly differ between groups? Does this agree
> with what you saw in the profile-of-means plot (Question 2.12)?

## Part E — Relationships between variables

### Mean vector, covariance and correlation

The **mean vector** is just the average of every variable. The **covariance
matrix** measures how pairs of variables vary together; the **correlation
matrix** is the same idea rescaled to lie between −1 and +1.

```{r cov-cor}
colMeans(hcv[, labs], na.rm = TRUE)                 # the mean vector
round(cov(hcv[, labs], use = "complete.obs"), 1)    # covariance matrix
round(cor(hcv[, labs], use = "complete.obs"), 2)    # correlation matrix
```

> **Question 2.14.** Look at the diagonal of the **covariance** matrix (each
> variable's variance). Creatinine's number is enormous compared with
> cholesterol's. Is creatinine "more variable" in a meaningful sense, or is this
> just because of its units and scale? Why is the **correlation** matrix easier
> to compare across tests?

### Which tests move together?

```{r cor-matrix, fig.width=7, fig.height=7}
cor_matrix <- cor(hcv[, labs], use = "complete.obs")
corrplot(cor_matrix, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.6, tl.col = "black",
         title = "Correlation between lab tests", mar = c(0,0,1,0))
```

```{r scatter}
plot(hcv$ALT, hcv$AST, pch = 19, col = rgb(0,0,1,0.3),
     main = "AST vs ALT", xlab = "ALT (U/L)", ylab = "AST (U/L)")
```

> **Question 2.15.** Which **pair** of lab tests is most strongly correlated? Does
> that match your clinical intuition? Now find a pair that is **negatively**
> correlated and try to explain why.

### A scatterplot of every pair at once

```{r pairs, fig.width=9, fig.height=9}
pairs(hcv[, c("ALB","AST","BIL","CHE","GGT","PROT")],
      col = c("#2166AC","#92C5DE","#F4A582","#B2182B")[hcv$Group],
      pch = 19, cex = 0.4, gap = 0.2,
      main = "Scatterplot matrix (coloured by diagnosis)")
```

> **Question 2.16.** In the scatterplot matrix, can you spot any pair of tests where
> the cirrhosis patients (dark red) clearly peel away from the donors (blue)?

## Part F — Pictures of whole patients

This classic plot draws **each patient as a small star**, so similar patients
look alike. We use a small, readable sample of a few patients per group.

```{r subset}
set.seed(1)
idx <- unlist(lapply(split(seq_len(nrow(hcv)), hcv$Group),
                     function(i) head(i[complete.cases(hcv[i, labs])], 3)))
sub <- hcv[idx, ]
sub_labs <- scale(sub[, labs])                       # standardised for fair shapes
rownames(sub_labs) <- paste(substr(sub$Group, 1, 4), seq_along(idx))
```

### Star (segment) plots

Each patient becomes a star: every ray is one lab test, longer = higher value.

```{r stars, fig.width=8, fig.height=6}
stars(sub_labs, draw.segments = TRUE, key.loc = c(ncol(sub_labs) + 4, 2),
      main = "Star plot: one star per patient", cex = 0.7)
```

> **Question 2.17.** In the star plot, do patients from the same diagnosis group
> look similar to each other? Point to two patients whose stars look almost the
> same, and two that look very different.

### A three-dimensional cloud

We can plot three tests at once and colour by diagnosis.

```{r scatter3d, fig.width=7, fig.height=6}
keep <- complete.cases(hcv[, c("AST","GGT","BIL")])
scatterplot3d(hcv$AST[keep], hcv$GGT[keep], hcv$BIL[keep],
              color = c("#2166AC","#92C5DE","#F4A582","#B2182B")[hcv$Group[keep]],
              pch = 19, xlab = "AST", ylab = "GGT", zlab = "BIL",
              main = "Three liver markers in 3-D")
```

> **Question 2.18.** In the 3-D cloud of AST, GGT and BIL, do the cirrhosis
> patients occupy a different region from the donors? Is it getting hard to "see"
> the separation with only three variables — and we have ten? (This is exactly the
> problem that PCA, in TP2, will solve.)

## Part G — Standardisation and wrap-up

We introduced standardisation in Part D (Question 2.12) and used it again for the
star plots. Here is the key check, and the reason it matters for everything that
follows — including PCA and clustering in TP2 and TP3.

```{r scale-check}
hcv_z <- scale(hcv[, labs])
round(colMeans(hcv_z, na.rm = TRUE), 2)        # all ~0
round(apply(hcv_z, 2, sd, na.rm = TRUE), 2)    # all 1
```

> **Question 2.19.** Before standardising, creatinine (values up to ~1000) would
> dominate any distance, average, or multivariate plot simply because its numbers
> are biggest. Why does putting all variables on the same scale make a *fairer*
> comparison between patients? Re-draw the correlation `corrplot` on `hcv_z` — does
> standardising change the correlations? (It should not — explain why.)

> **Question 2.20 (wrap-up).** In three or four sentences, summarise what you have
> learned: How big is each dataset? Which variables are skewed, and what fixes
> that? Which tests rise or fall with liver disease? Which tests move together?
> And which single tests best separate cirrhosis from healthy donors?

---

```{r session}
sessionInfo()
```
