---
title: "TP5 — Principal Component Analysis (PCA)"
subtitle: "Doctors as Data Scientists · Dimensionality Reduction and Patient Profiling"
author: "Massimo Andreatta"
date: "`r Sys.Date()`"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file, encoding=encoding,
  output_file=file.path(dirname(input_file), '..', out_dir, 'TP5_PCA_Hepatitis.html'))})
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, fig.align = "center")
```

> **How to use this notebook.** You do **not** need any programming experience.
> Read each short explanation, then press the green *play* button (▶) at the
> top-right of each grey *code chunk* to run it. Look at the output, and answer
> the **Questions** in the green boxes (write your answers in plain text under
> each one). The goal is to *understand the data and the method*, not to
> memorise code.

---

# Introduction to Principal Component Analysis

## The problem: too many variables, too few dimensions

In medicine we routinely measure many things about each patient — blood tests,
vital signs, imaging findings. When there are six, ten, or fifty measurements
per patient, it becomes impossible to *see* the data: you cannot draw a
50-dimensional scatter plot. Yet a lot of information in those measurements is
**redundant** — for example, two liver enzymes that always rise together carry
roughly the same information.

**Principal Component Analysis (PCA)** exploits this redundancy to compress
many variables into a small number of new variables — the **principal
components** — that still capture most of the interesting variation. Think of
it as finding the best angle from which to photograph a 3-D sculpture: you
lose one dimension, but if you choose the right angle you preserve the
important structure.

## What PCA actually does — step by step

Imagine you have measured two things about each patient: ejection fraction and
serum creatinine. You could plot every patient as a point in a 2-D scatter
plot. PCA asks: *what single axis through this cloud of points captures the
most spread?* That axis is **PC1**. A second axis, perpendicular to the first,
captures the next most spread — that is **PC2**. Together they form a new
coordinate system that is rotated relative to the original axes.

```{r pca-intuition, fig.width=7, fig.height=5, echo=FALSE}
set.seed(7)
n  <- 120
z1 <- rnorm(n); z2 <- rnorm(n, sd = 0.35)
theta <- -40 * pi / 180
x <- cos(theta) * z1 - sin(theta) * z2
y <- sin(theta) * z1 + cos(theta) * z2

pca_toy <- prcomp(cbind(x, y), scale. = FALSE)
v1 <- pca_toy$rotation[, 1]
v2 <- pca_toy$rotation[, 2]
sc <- 1.5

plot(x, y, pch = 19, col = adjustcolor("steelblue", alpha.f = 0.5), cex = 0.9,
     xlab = "Measurement 1  (e.g. ALT)",
     ylab = "Measurement 2  (e.g. AST)",
     main = "PCA: finding new axes through the data cloud",
     asp = 1)
arrows(0, 0, sc * v1[1], sc * v1[2],
       col = "red", lwd = 4, length = 0.15)
arrows(0, 0, sc * v2[1], sc * v2[2],
       col = "orange", lwd = 4, length = 0.15)
text(sc * v1[1] * 1.18, sc * v1[2] * 1.18, "PC1", col = "red",    cex = 1.2, font = 2)
text(sc * v2[1] * 1.45, sc * v2[2] * 1.45, "PC2", col = "orange", cex = 1.2, font = 2)
```

The red arrow (PC1) points in the direction of greatest spread — patients are
most spread out along this axis. The orange arrow (PC2) is perpendicular to
PC1 and captures the remaining spread.

In this toy example we still have two dimensions, so nothing is gained yet.
The real benefit appears when we have **many** variables (ten in our dataset):
PCA finds the directions of greatest variance in that ten-dimensional space,
and we can keep only the first two or three PCs — which together capture the
majority of the information — and discard the rest. We compress ten dimensions
into two while losing only a small fraction of information, making the data
possible to visualise and interpret.

## Key vocabulary

| Term | Plain-language meaning |
|---|---|
| **Principal component (PC)** | A new axis through the data; a weighted combination of the original variables |
| **Score** | A patient's coordinate on a PC — where they sit along that axis |
| **Loading** | How much each original variable contributes to a PC (its "weight" on that axis) |
| **Variance explained** | The % of total spread captured by a PC — larger means more informative |
| **Scree plot** | A bar chart of variance explained by each PC, used to decide how many to keep |
| **Biplot** | A combined plot showing both patient scores and variable loadings |

## What PCA can — and cannot — do

PCA is an **unsupervised** method: it looks only at the measurements, with no
knowledge of clinical outcomes. It will find the directions of greatest
variation, which may or may not align with what we care about clinically. When
the main sources of biological variation *do* align with disease severity or
prognosis, PCA makes that structure visible. When they do not (for instance,
if the dominant variation is technical noise), PCA can mislead. Always
interpret the components biologically before drawing conclusions.

---

# The dataset: hepatitis C blood panel

Each row is one person who had a blood sample analysed. The data come from the
[UCI HCV dataset](https://archive.ics.uci.edu/dataset/571/hcv+data), containing
healthy **blood donors** and patients with hepatitis C at increasing stages of
liver damage (**Hepatitis → Fibrosis → Cirrhosis**).

The ten numeric columns are routine **laboratory tests**:

| Code | Lab test | What it reflects |
|------|----------|------------------|
| ALB  | Albumin | Liver synthetic function (falls when liver fails) |
| ALP  | Alkaline phosphatase | Bile ducts / liver |
| ALT  | Alanine aminotransferase | Liver-cell damage (rises with inflammation) |
| AST  | Aspartate aminotransferase | Liver-cell damage |
| BIL  | Bilirubin | Liver clearance (rises in jaundice) |
| CHE  | Cholinesterase | Liver synthetic function (falls when liver fails) |
| CHOL | Cholesterol | Liver / metabolism |
| CREA | Creatinine | Kidney function |
| GGT  | Gamma-glutamyl transferase | Bile ducts / liver |
| PROT | Total protein | Overall protein level |

`Category` is the diagnosis and `Sex` is m/f.

This dataset is particularly well suited to PCA: the ten tests are not
independent (several measure related aspects of liver function), and disease
severity follows an ordered progression — an ideal setting for PCA to reveal
biological structure.

## Load the data

```{r load}
hcv <- read.csv("../data/hcvdat0.csv", row.names = 1)
head(hcv)
```

```{r setup-helpers}
# Create a readable diagnosis label ordered by severity
hcv$Group <- factor(gsub("^0.*", "Donor", hcv$Category),
                    levels = c("Donor", "1=Hepatitis", "2=Fibrosis", "3=Cirrhosis"),
                    labels = c("Donor", "Hepatitis", "Fibrosis", "Cirrhosis"))

labs <- c("ALB","ALP","ALT","AST","BIL","CHE","CHOL","CREA","GGT","PROT")

# Colour palette: blue (healthy) → red (severe)
grp_cols <- c(Donor     = "#2166AC",
              Hepatitis = "#92C5DE",
              Fibrosis  = "#F4A582",
              Cirrhosis = "#B2182B")
```

---

# Part A — Meet the data

## Size, groups, and missing values

```{r structure}
dim(hcv)
table(hcv$Group)
```

```{r missing}
colSums(is.na(hcv[, labs]))
cat("Patients with at least one missing lab value:", sum(!complete.cases(hcv[, labs])), "\n")
```

Some patients are missing one or more lab values. PCA requires complete data,
so we keep only patients with all ten measurements. We lose a small fraction of
the cohort but keep the analysis straightforward.

```{r complete}
hcv_c <- hcv[complete.cases(hcv[, labs]), ]
cat("Patients retained:", nrow(hcv_c), "\n")
table(hcv_c$Group)
```

> **Question 1.1.** How many patients are in the dataset in total, and how many
> remain after removing incomplete cases? Which disease group loses the most
> patients to missing values — and can you guess why certain tests might be
> missing more often than others?

## Distributions and skewness

Right-skewed variables distort PCA — extreme values in a few patients drag the
axes. Let us check all ten tests.

```{r histograms, fig.width=11, fig.height=7}
par(mfrow = c(2, 5), mar = c(3, 3, 2, 1))
for (v in labs) {
  hist(hcv_c[[v]], breaks = 35, col = "steelblue", border = NA,
       main = v, xlab = "")
}
par(mfrow = c(1, 1))
```

Several enzyme and bilirubin measurements (ALT, AST, BIL, GGT) are heavily
right-skewed — the majority of patients cluster at low values but a few reach
extreme highs. We apply a log-transformation to these before running PCA.

```{r log-transform}
hcv_c$log_ALT <- log(hcv_c$ALT)
hcv_c$log_AST <- log(hcv_c$AST)
hcv_c$log_BIL <- log(hcv_c$BIL)
hcv_c$log_GGT <- log(hcv_c$GGT)

labs_clean <- c("ALB","ALP","log_ALT","log_AST","log_BIL",
                "CHE","CHOL","CREA","log_GGT","PROT")
```

```{r log-check, fig.width=11, fig.height=3.5}
par(mfrow = c(1, 4))
for (v in c("log_ALT","log_AST","log_BIL","log_GGT")) {
  hist(hcv_c[[v]], breaks = 35, col = "darkorange", border = NA,
       main = v, xlab = "")
}
par(mfrow = c(1, 1))
```

> **Question 1.2.** Which of the ten tests are most heavily right-skewed in
> the original scale? Why might patients with cirrhosis drive the extreme
> values in ALT, AST and GGT specifically — what do these tests measure?

> **Question 1.3.** After log-transformation, do the distributions of ALT,
> AST, BIL and GGT look more symmetric? Why does a more symmetric distribution
> help PCA produce interpretable components?

---

# Part B — Why standardise before PCA?

PCA finds directions of **maximum variance**. A variable measured in large
numbers (e.g. creatinine in µmol/L) would dominate the result over one
measured in small numbers (e.g. cholesterol in mmol/L), even if both carry
equal biological information.

```{r raw-sd}
round(apply(hcv_c[, labs_clean], 2, sd, na.rm = TRUE), 3)
```

```{r scale}
X <- scale(hcv_c[, labs_clean])   # z-score: mean 0, sd 1

round(apply(X, 2, mean), 3)       # all ≈ 0
round(apply(X, 2, sd),   3)       # all = 1
```

> **Question 2.1.** Before scaling, which variable has the largest standard
> deviation? If we ran PCA without standardising, which variable would likely
> dominate PC1 — and would that be biologically meaningful, or just an artefact
> of units?

> **Question 2.2.** We already log-transformed four tests. Does log-
> transformation change the need for standardisation, or are both steps
> necessary for different reasons?

---

# Part C — Running PCA

## Computing the principal components

```{r pca}
pca <- prcomp(X, scale. = FALSE)   # already scaled
summary(pca)
```

## Scree plot

```{r scree, fig.width=6, fig.height=4}
var_exp <- pca$sdev^2 / sum(pca$sdev^2) * 100

barplot(var_exp,
        names.arg = paste0("PC", seq_along(var_exp)),
        col = "steelblue", border = NA,
        ylab = "Variance explained (%)",
        main = "Scree plot")
```

```{r cumvar, fig.width=6, fig.height=4}
plot(cumsum(var_exp), type = "b", pch = 19, col = "steelblue",
     xlab = "Number of PCs", ylab = "Cumulative variance explained (%)",
     main = "Cumulative variance explained", ylim = c(0, 100))
abline(h = 80, lty = 2, col = "grey60")
```

> **Question 3.1.** How much variance does PC1 explain on its own? Given that
> we have ten input variables, is it surprising that a single axis can capture
> such a large fraction? What does this suggest about the redundancy among
> these liver tests?

> **Question 3.2.** How many PCs would you need to retain to explain at least
> 80 % of the variance? Where is the elbow in the scree plot?

---

# Part D — Patients in PCA space

## The score plot coloured by diagnosis

Unlike in many datasets, here we **know** the true diagnosis of every patient.
We colour the score plot by disease group to see whether PCA — which had no
knowledge of the diagnosis — recovers the biological structure.

```{r scores}
scores <- as.data.frame(pca$x)
scores$Group <- hcv_c$Group
col_pts <- grp_cols[as.character(scores$Group)]
```

```{r score-plot, fig.width=7, fig.height=5.5}
plot(scores$PC1, scores$PC2,
     col = col_pts, pch = 19, cex = 0.7,
     xlab = paste0("PC1 (", round(var_exp[1], 1), " %)"),
     ylab = paste0("PC2 (", round(var_exp[2], 1), " %)"),
     main = "PCA score plot — coloured by diagnosis")
legend("topright", legend = names(grp_cols), col = grp_cols,
       pch = 19, cex = 0.85, pt.cex = 1.1)
```

> **Question 4.1.** Do the four diagnosis groups separate along PC1, PC2, or
> both? Is the ordering of the groups along PC1 consistent with the biological
> progression of liver disease (Donor → Hepatitis → Fibrosis → Cirrhosis)?

> **Question 4.2.** Are the Donor and Cirrhosis groups well separated, or do
> they overlap substantially? What about Hepatitis and Fibrosis — are they
> easier or harder to distinguish? Does this reflect the clinical reality of
> diagnosing intermediate stages of liver disease?

## Mapping individual lab tests onto the PCA

```{r color-helper}
map_color <- function(x, n = 100) {
  pal <- colorRampPalette(c("red", "white", "blue"))(n)
  idx <- cut(x, breaks = n, labels = FALSE, include.lowest = TRUE)
  pal[idx]
}
```

We plot the same patient cloud coloured by selected lab values to see which
measurements explain the position along each PC axis.

```{r pca-labs-damage, fig.width=11, fig.height=4}
par(mfrow = c(1, 3))

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.6,
     col = map_color(hcv_c$log_ALT),
     main = "log(ALT) — liver-cell damage",
     xlab = "PC1", ylab = "PC2")

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.6,
     col = map_color(hcv_c$log_GGT),
     main = "log(GGT) — bile duct / liver",
     xlab = "PC1", ylab = "PC2")

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.6,
     col = map_color(hcv_c$log_BIL),
     main = "log(BIL) — bilirubin / jaundice",
     xlab = "PC1", ylab = "PC2")

par(mfrow = c(1, 1))
```

```{r pca-labs-synthetic, fig.width=8, fig.height=4}
par(mfrow = c(1, 2))

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.6,
     col = map_color(hcv_c$ALB),
     main = "ALB — liver synthetic function",
     xlab = "PC1", ylab = "PC2")

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.6,
     col = map_color(hcv_c$CHE),
     main = "CHE — liver synthetic function",
     xlab = "PC1", ylab = "PC2")

par(mfrow = c(1, 1))
```

*In all plots: red = low values, blue = high values.*

> **Question 4.3.** ALT, GGT and BIL **rise** with liver damage, while ALB
> and CHE **fall** as the liver loses its synthetic function. Looking at the
> colour gradients: do the "rise" tests (high = blue) and the "fall" tests
> (high = blue) sit on the *same* side of PC1, or on *opposite* sides? What
> does this tell you about the biological meaning of PC1?

---

# Part E — What drives each PC? Loadings

The **loadings** reveal how much each lab test contributes to each PC and in
which direction.

```{r loadings}
round(pca$rotation[, 1:4], 3)
```

```{r loading-bars, fig.width=9, fig.height=4}
par(mfrow = c(1, 2))

barplot(sort(pca$rotation[, 1]),
        horiz = TRUE, las = 1, col = "steelblue", border = NA,
        main = "PC1 loadings", xlab = "Loading value")
abline(v = 0, col = "grey40")

barplot(sort(pca$rotation[, 2]),
        horiz = TRUE, las = 1, col = "darkorange", border = NA,
        main = "PC2 loadings", xlab = "Loading value")
abline(v = 0, col = "grey40")

par(mfrow = c(1, 1))
```

## Biplot

```{r biplot, fig.width=7, fig.height=6}
biplot(pca, scale = 0, cex = c(0.4, 0.9),
       col = c("grey70", "tomato"),
       main = "PCA biplot — patients and variable loadings")
```

> **Question 5.1.** Which lab tests have the strongest loadings on PC1 — and
> do the tests that *rise* with liver damage (ALT, AST, GGT, BIL) load in the
> *opposite* direction from those that *fall* (ALB, CHE)? Does this confirm
> your interpretation from Question 4.3?

> **Question 5.2.** In the biplot, arrows pointing in the same direction
> indicate correlated variables. Which pairs of tests are most strongly
> correlated? Does this make biological sense — for example, are ALT and AST
> (both markers of liver-cell damage) in a similar direction?

> **Question 5.3.** Which variables contribute most to PC2? Based on those
> loadings, what aspect of patient variation might PC2 be capturing — and why
> might it be independent of the main liver-disease axis?

---

# Part F — Unsupervised clustering

With the HCV dataset we can do something that is rarely possible in real
practice: we know the true diagnosis, so we can directly evaluate how well
**unsupervised** clustering (which ignores the labels) recovers the biological
groups.

## Elbow plot

```{r elbow, fig.width=6, fig.height=4}
set.seed(42)
npc <- 4
data <- pca$x[,1:npc]

wss <- sapply(1:8, function(k) {
  kmeans(data, centers = k, nstart = 25)$tot.withinss
})

plot(1:8, wss, type = "b", pch = 19, col = "steelblue",
     xlab = "Number of clusters (k)",
     ylab = "Total within-cluster sum of squares",
     main = "Elbow plot for k-means")
```

> **Question 6.1.** We have four known disease groups. Does the elbow plot
> suggest *k* = 4 as a natural number of clusters, or does the curve flatten
> earlier? What does this tell us about whether the four clinical groups
> form truly distinct clusters in the measurement space?

## Clustering with k = 2

```{r kmeans}
set.seed(42)
k = 2

data <- pca$x[,1:npc]
km   <- kmeans(data, centers = k, nstart = 25)
scores$cluster <- as.factor(km$cluster)
```

```{r cluster-plot, fig.width=7, fig.height=5.5}
cl_cols <- c("1" = "#1b9e77", "2" = "#d95f02")

plot(scores$PC1, scores$PC2,
     col = cl_cols[scores$cluster], pch = 19, cex = 0.7,
     xlab = paste0("PC1 (", round(var_exp[1], 1), " %)"),
     ylab = paste0("PC2 (", round(var_exp[2], 1), " %)"),
     main = "K-means clusters (k = 2) in PCA space")
legend("topright", legend = paste("Cluster", 1:2),
       col = cl_cols, pch = 19, cex = 0.85)
```

## How well do clusters match the known diagnoses?

```{r confusion}
tab <- table(Cluster = scores$cluster, Diagnosis = scores$Group)
tab
```

```{r confusion-pct}
round(prop.table(tab, margin = 1) * 100, 1)
```

```{r cluster-bar, fig.width=7, fig.height=4.5}
barplot(t(prop.table(tab, margin = 1)) * 100,
        col  = grp_cols, border = NA,
        legend.text = names(grp_cols),
        args.legend = list(x = "topright", cex = 0.85),
        ylab = "Patients (%)", xlab = "Cluster",
        main = "Diagnosis composition of each k-means cluster")
```

> **Question 6.2.** Are any clusters highly enriched for a single diagnosis
> (e.g. one cluster = mostly Cirrhosis)? Which clinical groups are most
> cleanly separated, and which tend to be mixed together across clusters?

> **Question 6.3.** The clustering algorithm had no access to the diagnosis
> labels — it worked only from the ten lab values. Yet it partially recovers
> the clinical groups. What does this tell us about the information content of
> routine blood tests for staging liver disease?

---

# Part G — Wrap-up

```{r session}
sessionInfo()
```

> **Question 7.1 (wrap-up).** In four or five sentences, summarise the
> journey: How much of the variation in ten lab tests can be captured by just
> two or three axes? What do those axes represent biologically? Does the
> disease progression (Donor → Hepatitis → Fibrosis → Cirrhosis) appear as a
> gradient in PCA space, and which specific tests drive that gradient? What
> does the clustering analysis add to this picture?
