---
title: "TP5 — Principal Component Analysis (PCA)"
subtitle: "Doctors as Data Scientists · Molecular Subtypes of Breast Cancer"
author: "Massimo Andreatta"
date: "`r Sys.Date()`"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file, encoding=encoding,
  output_file=file.path(dirname(input_file), '..', out_dir, 'TP5_PCA_BRCA_PAM50.html'))})
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, fig.align = "center")
```

> **How to use this notebook.** You do **not** need any programming experience.
> Read each short explanation, then press the green *play* button (▶) at the
> top-right of each grey *code chunk* to run it. Look at the output, and answer
> the **Questions** in the green boxes (write your answers in plain text under
> each one). The goal is to *understand the data and the method*, not to
> memorise code.

---

# Introduction to Principal Component Analysis

## The problem: too many variables, too few dimensions

Modern medicine no longer measures only a handful of things per patient.
In oncology, it is now possible to measure the activity level of **every
gene in a tumour** — typically more than 20,000 genes — from a single biopsy.
This wealth of information is extraordinary, but it creates a problem:
you cannot draw a 20,000-dimensional scatter plot. Even with only 50 genes
you cannot visualise all the data at once.

Yet a lot of information in those measurements is **redundant**: genes in
the same biological pathway tend to rise and fall together. **Principal
Component Analysis (PCA)** exploits this redundancy to compress many
variables into a small number of new variables — the **principal
components** — that still capture most of the interesting variation.

## What PCA actually does — step by step

Imagine you have measured the activity of two genes in each tumour sample.
You could plot every sample as a point in a 2-D scatter plot. PCA asks:
*what single axis through this cloud of points captures the most spread?*
That axis is **PC1**. A second axis, perpendicular to the first, captures
the next most spread — that is **PC2**. Together they form a new coordinate
system rotated relative to the original axes.

```{r pca-intuition, fig.width=7, fig.height=5, echo=FALSE}
set.seed(7)
n  <- 120
z1 <- rnorm(n); z2 <- rnorm(n, sd = 0.35)
theta <- -40 * pi / 180
x <- cos(theta) * z1 - sin(theta) * z2
y <- sin(theta) * z1 + cos(theta) * z2

pca_toy <- prcomp(cbind(x, y), scale. = FALSE)
v1 <- pca_toy$rotation[, 1]
v2 <- pca_toy$rotation[, 2]
sc <- 1.5

plot(x, y, pch = 19, col = adjustcolor("steelblue", alpha.f = 0.5), cex = 0.9,
     xlab = "Gene 1 expression  (e.g. ESR1)",
     ylab = "Gene 2 expression  (e.g. KRT5)",
     main = "PCA: finding new axes through the data cloud",
     asp = 1)
arrows(0, 0, sc * v1[1], sc * v1[2],
       col = "red", lwd = 4, length = 0.15)
arrows(0, 0, sc * v2[1], sc * v2[2],
       col = "orange", lwd = 4, length = 0.15)
text(sc * v1[1] * 1.18, sc * v1[2] * 1.18, "PC1", col = "red",    cex = 1.2, font = 2)
text(sc * v2[1] * 1.45, sc * v2[2] * 1.45, "PC2", col = "orange", cex = 1.2, font = 2)
```

The red arrow (PC1) points in the direction of greatest spread — tumour
samples are most spread out along this axis. The orange arrow (PC2) is
perpendicular to PC1 and captures the remaining spread.

The real benefit appears when we have **many** variables. PCA finds the
directions of greatest variance in the N-dimensional space and we can keep
only the first two or three PCs — which together capture the majority of
the information — and discard the rest. We compress 50 dimensions into two
while losing only a small fraction of information, making the data possible
to visualise and interpret.

## Key vocabulary

| Term | Plain-language meaning |
|---|---|
| **Principal component (PC)** | A new axis through the data; a weighted combination of the original variables |
| **Score** | A tumour's coordinate on a PC — where it sits along that axis |
| **Loading** | How much each original gene contributes to a PC (its "weight" on that axis) |
| **Variance explained** | The % of total spread captured by a PC — larger means more informative |
| **Scree plot** | A bar chart of variance explained by each PC, used to decide how many to keep |
| **Biplot** | A combined plot showing both tumour scores and gene loadings |

## What PCA can — and cannot — do

PCA is an **unsupervised** method: it looks only at the measurements, with
no knowledge of clinical labels. It will find the directions of greatest
variation, which may or may not align with what we care about clinically.
When the main sources of biological variation *do* align with cancer
subtype or prognosis, PCA makes that structure visible. Always interpret
the components biologically before drawing conclusions.

---

# The dataset: breast cancer gene expression (TCGA-BRCA)

## Background: breast cancer is not one disease

For decades, breast cancer was treated largely as a single disease. In the
early 2000s, gene-expression profiling revealed that breast cancer
comprises at least **four molecularly distinct subtypes** with very
different biology, prognosis, and treatment response:

| Subtype | Key biology | Hormone receptors | Prognosis |
|---|---|---|---|
| **Luminal A** (LumA) | ER/PR-positive, low proliferation | ER+, PR+ | Best |
| **Luminal B** (LumB) | ER/PR-positive, higher proliferation | ER+/−, PR+/− | Intermediate |
| **HER2-enriched** (Her2) | HER2 gene amplification | HER2+, often ER− | Variable |
| **Basal-like** (Basal) | Triple-negative, high genomic instability | ER−, PR−, HER2− | Worst |
| **Normal-like** | Gene pattern resembles normal breast tissue | Variable | Intermediate |

The **PAM50** test (Parker *et al.* 2009, *J Clin Oncol*) is a clinical
assay that measures 50 carefully selected genes and assigns each tumour to
one of these subtypes. It is used clinically to guide treatment decisions
and predict the benefit of chemotherapy.

## The 50 PAM50 genes and what they measure

The 50 genes span several biological programmes:

| Biological role | Key genes |
|---|---|
| Oestrogen / hormone signalling | *ESR1*, *PGR*, *NAT1*, *FOXA1*, *BCL2*, *SLC39A6* |
| HER2 / growth factor signalling | *ERBB2*, *GRB7*, *FGFR4*, *MMP11* |
| Basal / myoepithelial markers | *KRT5*, *KRT14*, *KRT17*, *EGFR*, *CDH3*, *FOXC1* |
| Cell proliferation | *MKI67*, *CCNB1*, *CCNE1*, *CDC6*, *CDC20*, *RRM2*, *BIRC5* |
| Other structural / signalling | *MDM2*, *MYC*, *MELK*, *CENPF*, … |

Expression values are on a **log₂ scale** (log₂ of normalised RNA-seq
counts). All 50 genes are already on the same numeric scale, so no further
log-transformation is needed. We will still need to **standardise** them
before PCA (see Part B).

## The data

The dataset contains **522 primary breast cancer samples** from The Cancer
Genome Atlas (TCGA-BRCA), each with:
- Expression of the 50 PAM50 genes (continuous, log₂-normalised)
- PAM50 subtype label (`pam50`)
- Clinical variables: oestrogen receptor status (`er_status`), HER2 status
  (`her2_status`), progesterone receptor status (`pr_status`), age, and
  disease stage

## Load the data

```{r load}
brca <- read.csv("../data/brca_pam50.csv", row.names = 1)
head(brca[, 1:8])   # show first few columns
```

```{r setup-labels}
# Gene expression columns
gene_cols <- colnames(brca)[7:56]

# Colour palette for PAM50 subtypes
subtype_cols <- c(
  LumA       = "#1f77b4",   # blue — best prognosis
  LumB       = "#aec7e8",   # light blue
  Her2       = "#ff7f0e",   # orange
  Basal      = "#d62728",   # red — worst prognosis
  NormalLike = "#7f7f7f"    # grey
)
```

---

# Part A — Meet the data

## Size and subtype distribution

```{r structure}
dim(brca)
cat("Samples per PAM50 subtype:\n")
table(brca$pam50)
```

```{r subtype-bar, fig.width=6, fig.height=4}
counts <- table(brca$pam50)
barplot(counts[c("LumA","LumB","Her2","Basal","NormalLike")],
        col   = subtype_cols[c("LumA","LumB","Her2","Basal","NormalLike")],
        border = NA,
        ylab  = "Number of tumours",
        main  = "PAM50 subtype distribution (TCGA-BRCA)")
```

> **Question 1.1.** How many tumour samples are in the dataset? Which is
> the most common PAM50 subtype, and which is the rarest? Does the
> prevalence of Luminal A tumours in this dataset surprise you, given that
> ER-positive breast cancer is the most common type diagnosed clinically?

## Hormone receptor and HER2 status

The PAM50 molecular subtypes were derived from gene expression, but they
should correspond closely to the clinical receptor status measured in
routine pathology.

```{r receptor-table}
cat("ER status by PAM50 subtype:\n")
print(table(brca$pam50, brca$er_status))

cat("\nHER2 status by PAM50 subtype:\n")
print(table(brca$pam50, brca$her2_status))
```

```{r receptor-plots, fig.width=10, fig.height=4.5}
subtype_order <- c("LumA", "LumB", "Her2", "Basal", "NormalLike")

# ER status — keep Positive / Negative; collapse rare categories
er_clean  <- ifelse(brca$er_status  %in% c("Positive","Negative"), brca$er_status,  NA)
her2_clean <- ifelse(brca$her2_status %in% c("Positive","Negative"), brca$her2_status, NA)

tab_er   <- table(factor(brca$pam50, levels = subtype_order),
                  factor(er_clean,   levels = c("Positive","Negative")))
tab_her2 <- table(factor(brca$pam50, levels = subtype_order),
                  factor(her2_clean, levels = c("Positive","Negative")))

er_pct   <- prop.table(tab_er,   margin = 1) * 100
her2_pct <- prop.table(tab_her2, margin = 1) * 100

er_cols   <- c(Positive = "#e41a1c", Negative = "#bdbdbd")
her2_cols <- c(Positive = "#ff7f00", Negative = "#bdbdbd")

par(mfrow = c(1, 2))

barplot(t(er_pct),
        col    = er_cols, border = NA,
        beside = FALSE,
        legend.text = c("ER+", "ER−"),
        args.legend = list(x = "topright", cex = 0.9,
                           fill = er_cols, border = NA),
        ylab = "Tumours (%)", xlab = "PAM50 subtype",
        main = "ER status by molecular subtype",
        ylim = c(0, 115))

barplot(t(her2_pct),
        col    = her2_cols, border = NA,
        beside = FALSE,
        legend.text = c("HER2+", "HER2−"),
        args.legend = list(x = "topright", cex = 0.9,
                           fill = her2_cols, border = NA),
        ylab = "Tumours (%)", xlab = "PAM50 subtype",
        main = "HER2 status by molecular subtype",
        ylim = c(0, 115))

par(mfrow = c(1, 1))
```

> **Question 1.2.** Are Luminal A and B tumours predominantly ER-positive?
> Are HER2-enriched tumours predominantly HER2-positive by IHC/FISH? Are
> Basal-like tumours predominantly ER-negative and HER2-negative (i.e.
> triple-negative)? What does this correspondence — or any discordance —
> tell you about the relationship between the molecular subtype (RNA-based)
> and routine pathology testing (protein-based)?

## Distributions of key gene markers

Let us look at the distribution of a few biologically informative genes
across the dataset.

```{r key-gene-hist, fig.width=10, fig.height=4}
key_genes <- c("ESR1", "ERBB2", "KRT5", "MKI67")
par(mfrow = c(1, 4), mar = c(4, 3, 3, 1))
for (g in key_genes) {
  hist(brca[[g]], breaks = 30, col = "steelblue", border = NA,
       main = g, xlab = "log₂ expression")
}
par(mfrow = c(1, 1))
```

*ESR1* encodes the oestrogen receptor; *ERBB2* encodes HER2; *KRT5* is a
basal cytokeratin; *MKI67* (Ki-67) measures cell proliferation.

> **Question 1.3.** Do any of these four genes show a **bimodal**
> distribution (two peaks rather than one)? What would a bimodal
> distribution suggest about the tumour population — could it reflect a
> mixture of two molecularly distinct groups?

## Gene–gene correlation

Genes in the same biological pathway tend to be co-expressed. We can
visualise this with a correlation heatmap.

```{r corr-heatmap, fig.width=8, fig.height=7}
cor_mat  <- cor(brca[, gene_cols], use = "complete.obs")
col_heat <- colorRampPalette(c("#4393c3", "white", "#d6604d"))(100)

# Hierarchical clustering to reorder genes so co-expressed blocks appear together
gene_dist <- as.dist(1 - cor_mat)
gene_clust <- hclust(gene_dist, method = "average")
gene_order <- gene_clust$order

heatmap(cor_mat[gene_order, gene_order],
        Rowv = NA, Colv = NA,      # order already set; suppress re-clustering
        col  = col_heat,
        scale = "none",
        margins = c(6, 6),
        cexRow = 0.6, cexCol = 0.6,
        main = "Gene–gene correlation (50 PAM50 genes, clustered)")
```

> **Question 1.4.** After hierarchical clustering, genes with similar
> expression patterns across tumours are grouped together and appear as
> red blocks along the diagonal. Can you identify at least two distinct
> blocks? (Hint: look for a cluster containing *ESR1*, *PGR*, *FOXA1*
> and another containing *MKI67*, *CCNB1*, *RRM2*.) What biological
> programme does each block represent, and why would genes in the same
> programme tend to be co-expressed?

---

# Part B — Why standardise before PCA?

PCA finds the directions of **maximum variance**. If one gene has a much
larger spread than another, it will dominate the PCA result simply because
of its larger variance — not because it is more biologically informative.

```{r raw-sd}
gene_sds <- apply(brca[, gene_cols], 2, sd, na.rm = TRUE)
cat("Standard deviation of each gene (before scaling):\n")
print(round(sort(gene_sds, decreasing = TRUE), 2))
```

```{r sd-barplot, fig.width=9, fig.height=4}
barplot(sort(gene_sds, decreasing = TRUE),
        las = 2, cex.names = 0.6,
        col = "steelblue", border = NA,
        ylab = "Standard deviation",
        main = "Gene expression variability across 522 tumours")
```

```{r scale}
X <- scale(brca[, gene_cols])    # standardise to mean 0, sd 1
round(apply(X, 2, mean), 3)[1:5] # confirm: first 5 means ≈ 0
round(apply(X, 2, sd),   3)[1:5] # confirm: first 5 sds = 1
```

After standardisation every gene contributes on equal footing: we compare
*patterns of variation across tumours*, not raw expression magnitudes.

> **Question 2.1.** Before scaling, which gene has the largest standard
> deviation? Would it be fair to let this gene dominate PC1 just because
> it happens to vary more than others — or should every gene have equal
> weight in finding the major axes of biological variation?

> **Question 2.2.** Note that the data are already on a log₂ scale
> (range ≈ 0–18), so the *units* are already harmonised. Yet the standard
> deviations still differ by a factor of ~5 across genes. Does this mean
> log-transformation is enough on its own, or is standardisation a
> separate and necessary step?

---

# Part C — Running PCA

## Computing the principal components

`prcomp()` rotates the data so that PC1 captures as much variance as
possible, PC2 captures as much of the *remaining* variance as possible,
and so on. With 50 input genes, we get 50 principal components in total.

```{r pca}
pca <- prcomp(X, scale. = FALSE)   # already scaled above
summary(pca)$importance[, 1:8]     # show first 8 PCs
```

## Scree plot — how many PCs matter?

```{r scree, fig.width=7, fig.height=4}
var_exp <- pca$sdev^2 / sum(pca$sdev^2) * 100

barplot(var_exp[1:15],
        names.arg = paste0("PC", 1:15),
        col = "steelblue", border = NA,
        ylab = "Variance explained (%)",
        main = "Scree plot (first 15 PCs)")
```

```{r cumvar, fig.width=6, fig.height=4}
plot(cumsum(var_exp), type = "b", pch = 19, col = "steelblue",
     xlab = "Number of PCs", ylab = "Cumulative variance explained (%)",
     main = "Cumulative variance explained", ylim = c(0, 100))
abline(h = 80, lty = 2, col = "grey60")
```

> **Question 3.1.** How much variance (%) does PC1 alone explain? Given
> that we started with 50 genes, is it surprising that a single axis
> can capture such a large fraction of the total variation? What does
> this suggest about the redundancy among the 50 PAM50 genes?

> **Question 3.2.** How many PCs are needed to capture at least 80 % of
> the total variance? Where does the scree plot show an obvious "elbow"?
> What does this suggest about how many truly independent sources of
> variation exist in these 50 genes?

---

# Part D — Tumours in PCA space

## The score plot coloured by PAM50 subtype

Each tumour becomes a **single point** in PCA space. Tumours close
together have similar gene expression profiles; tumours far apart are
molecularly distinct. Crucially, the PCA was computed with **no knowledge
of the subtype labels** — we now colour the points by subtype to see
whether PCA rediscovers the biological structure.

```{r scores}
scores <- as.data.frame(pca$x)
scores$subtype <- factor(brca$pam50,
                         levels = c("LumA","LumB","Her2","Basal","NormalLike"))
col_pts <- subtype_cols[as.character(scores$subtype)]
```

```{r score-plot, fig.width=7, fig.height=5.5}
plot(scores$PC1, scores$PC2,
     col  = col_pts, pch = 19, cex = 0.7,
     xlab = paste0("PC1 (", round(var_exp[1], 1), " %)"),
     ylab = paste0("PC2 (", round(var_exp[2], 1), " %)"),
     main = "PCA score plot — coloured by PAM50 subtype")
legend("topright",
       legend = names(subtype_cols),
       col    = subtype_cols,
       pch = 19, cex = 0.85, pt.cex = 1.1)
```

> **Question 4.1.** Do the five PAM50 subtypes separate in PCA space?
> Which subtypes occupy the most distinct, non-overlapping regions? Which
> subtypes show the most overlap? Is this consistent with the clinical
> observation that Luminal A and Luminal B tumours are biologically more
> similar to each other than to Basal-like tumours?

> **Question 4.2.** Along which axis (PC1 or PC2, or both) do the Luminal
> and Basal tumours separate most clearly? Where do HER2-enriched tumours
> sit relative to these two poles? Does the arrangement suggest that the
> major axis of variation in breast cancer gene expression corresponds to
> the ER+/ER− distinction?

## Mapping individual genes onto the PCA

We colour the same score plot by the expression of specific marker genes
to reveal which genes explain tumour positions along each PC axis.

```{r color-helper}
map_color <- function(x, n = 100) {
  pal <- colorRampPalette(c("navy", "white", "firebrick"))(n)
  idx <- cut(x, breaks = n, labels = FALSE, include.lowest = TRUE)
  pal[idx]
}
```

### Hormone-receptor genes

```{r pca-hr, fig.width=11, fig.height=4}
par(mfrow = c(1, 3))

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.6,
     col  = map_color(brca$ESR1),
     main = "ESR1 (oestrogen receptor gene)",
     xlab = "PC1", ylab = "PC2")

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.6,
     col  = map_color(brca$PGR),
     main = "PGR (progesterone receptor gene)",
     xlab = "PC1", ylab = "PC2")

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.6,
     col  = map_color(brca$FOXA1),
     main = "FOXA1 (luminal transcription factor)",
     xlab = "PC1", ylab = "PC2")

par(mfrow = c(1, 1))
```

### Basal and HER2 marker genes

```{r pca-basal-her2, fig.width=11, fig.height=4}
par(mfrow = c(1, 3))

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.6,
     col  = map_color(brca$KRT5),
     main = "KRT5 (basal cytokeratin)",
     xlab = "PC1", ylab = "PC2")

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.6,
     col  = map_color(brca$ERBB2),
     main = "ERBB2 (HER2 gene)",
     xlab = "PC1", ylab = "PC2")

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.6,
     col  = map_color(brca$MKI67),
     main = "MKI67 (proliferation marker, Ki-67)",
     xlab = "PC1", ylab = "PC2")

par(mfrow = c(1, 1))
```

*In all plots: navy = low expression, red = high expression.*

> **Question 4.3.** Looking at the ESR1 and KRT5 plots: do tumours with
> **high ESR1** (red) sit on the same side of PC1 as tumours with **high
> KRT5** (red), or on *opposite* sides? ESR1 is the gene encoding the
> oestrogen receptor, which drives growth in Luminal tumours; KRT5 is a
> structural protein expressed in basal/myoepithelial cells. Does the
> geometry of the PCA confirm that these two programmes are biologically
> opposed?

> **Question 4.4.** Where do tumours with high *ERBB2* expression sit in
> PCA space? Is ERBB2 expression primarily driving the PC1 axis (the
> main Luminal vs Basal axis), or does it contribute more to a different
> direction? Based on what you know about HER2-positive breast cancer,
> would you expect HER2-enriched tumours to cluster with Luminal or with
> Basal-like tumours?

> **Question 4.5.** *MKI67* (which encodes Ki-67, a widely used clinical
> proliferation marker) is elevated in which region of PCA space? Is
> proliferation higher in Luminal A or Luminal B tumours? This mirrors a
> key clinical difference between the two Luminal subtypes — Luminal B is
> more proliferative and carries a worse prognosis.

---

# Part E — What drives each PC? Loadings

The **loadings** (in `pca$rotation`) tell us how much each gene contributes
to each PC. A large positive loading means that tumours with high expression
of that gene sit at the *positive* end of the PC; a large negative loading
places them at the *negative* end.

```{r loadings-table}
# Top 10 genes by absolute loading on PC1 and PC2
top_pc1 <- sort(pca$rotation[, 1])[c(1:5, 46:50)]   # 5 most negative, 5 most positive
top_pc2 <- sort(pca$rotation[, 2])[c(1:5, 46:50)]
cat("Top genes on PC1:\n"); print(round(top_pc1, 3))
cat("\nTop genes on PC2:\n"); print(round(top_pc2, 3))
```

```{r loading-bars, fig.width=10, fig.height=5}
n_show <- 10   # top 5 positive + top 5 negative

pick_top <- function(loadings, n) {
  s <- sort(loadings)
  c(head(s, n %/% 2), tail(s, n %/% 2))
}

par(mfrow = c(1, 2))

top1 <- pick_top(pca$rotation[, 1], n_show)
barplot(top1, horiz = TRUE, las = 1, cex.names = 0.85,
        col = ifelse(top1 > 0, "steelblue", "tomato"), border = NA,
        main = "PC1 — top 10 genes", xlab = "Loading value")
abline(v = 0, col = "grey40")

top2 <- pick_top(pca$rotation[, 2], n_show)
barplot(top2, horiz = TRUE, las = 1, cex.names = 0.85,
        col = ifelse(top2 > 0, "darkorange", "tomato"), border = NA,
        main = "PC2 — top 10 genes", xlab = "Loading value")
abline(v = 0, col = "grey40")

par(mfrow = c(1, 1))
```

## Biplot — tumours and key genes together

The **biplot** overlays tumour positions (dots) with arrows for a hand-picked
panel of biologically important genes. Arrows pointing in the same direction
indicate genes that increase together; opposite arrows indicate genes that move
in opposite directions across tumours.

```{r biplot, fig.width=8, fig.height=7}
# Hand-picked panel covering the main biological axes
panel_genes <- c("ESR1", "PGR", "FOXA1", "NAT1",   # hormone / luminal
                 "KRT5", "KRT14", "EGFR",            # basal
                 "ERBB2", "GRB7",                    # HER2 amplicon
                 "MKI67")                            # proliferation

pca_sub          <- pca
pca_sub$rotation <- pca$rotation[panel_genes, , drop = FALSE]

# xlabs = "." replaces each sample label with a dot
biplot(pca_sub, scale = 0,
       xlabs = rep(".", nrow(pca$x)),
       cex   = c(3.0, 0.9),
       col   = c("grey50", "tomato"),
       main  = "Biplot — PAM50 tumours and key gene loadings")
```

> **Question 5.1.** Looking at the PC1 loading barplot: which genes have
> the strongest **positive** loadings, and which have the strongest
> **negative** loadings? Based on their known biology, propose a plain-
> English interpretation for PC1: is it primarily an axis of *hormone
> signalling*, *basal identity*, or a contrast between the two?

> **Question 5.2.** In the biplot, do the arrows for hormone-receptor
> genes (*ESR1*, *PGR*, *FOXA1*) and basal markers (*KRT5*, *KRT14*,
> *KRT17*) point in roughly *opposite* directions? Luminal tumours are
> ER-positive with high hormone signalling; Basal-like tumours are
> ER-negative with high expression of basal keratins. Does the biplot
> geometry reflect this biological opposition?

> **Question 5.3.** Which genes load most strongly on PC2? Based on those
> genes, what biological programme might PC2 be capturing? (Hint: look at
> *ERBB2*, *GRB7*, and *MKI67* — the HER2 amplicon genes and the
> proliferation marker.)

---

# Part F — Unsupervised clustering of tumours

We know the PAM50 subtype label for every tumour. Now we ask: can we
**rediscover** these subtypes **without** using the labels — only from the
50 gene expression values?

**K-means clustering** assigns each tumour to one of *k* groups so that
tumours in the same group are as similar as possible in PCA space. Because
we already know the true subtypes, we can directly evaluate how well
unsupervised clustering matches the biology.

## Choosing the number of clusters (elbow plot)

```{r elbow, fig.width=6, fig.height=4}
set.seed(42)
npc <- 5    # use first 5 PCs for clustering

wss <- sapply(1:10, function(k) {
  kmeans(pca$x[, 1:npc], centers = k, nstart = 25)$tot.withinss
})

plot(1:10, wss, type = "b", pch = 19, col = "steelblue",
     xlab = "Number of clusters (k)",
     ylab = "Total within-cluster sum of squares",
     main = "Elbow plot for k-means")
```

> **Question 6.1.** There are five known PAM50 subtypes (though
> Normal-like has only 8 samples). Does the elbow plot suggest *k* = 5 as
> a clear inflection point, or does the curve flatten earlier? What might
> this tell us about whether all five subtypes form truly distinct clusters
> in 50-gene expression space?

## Clustering with k = 4

We choose *k* = 4 to match the four major clinical subtypes (LumA, LumB,
Her2, Basal), setting aside the rare Normal-like category.

```{r kmeans}
set.seed(42)
k <- 4
km <- kmeans(pca$x[, 1:npc], centers = k, nstart = 25)
scores$cluster <- as.factor(km$cluster)
```

```{r cluster-plot, fig.width=7, fig.height=5.5}
cl_cols <- c("1" = "#e41a1c", "2" = "#377eb8",
             "3" = "#4daf4a", "4" = "#984ea3")

plot(scores$PC1, scores$PC2,
     col = cl_cols[scores$cluster], pch = 19, cex = 0.7,
     xlab = paste0("PC1 (", round(var_exp[1], 1), " %)"),
     ylab = paste0("PC2 (", round(var_exp[2], 1), " %)"),
     main = "K-means clusters (k = 4) in PCA space")
legend("topright", legend = paste("Cluster", 1:4),
       col = cl_cols, pch = 19, cex = 0.85)
```

## How well do the unsupervised clusters match PAM50 subtypes?

```{r confusion}
tab <- table(Cluster = scores$cluster, PAM50 = scores$subtype)
tab
```

```{r confusion-pct}
cat("Row percentages (% of each cluster made up of each subtype):\n")
round(prop.table(tab, margin = 1) * 100, 1)
```

```{r cluster-bar, fig.width=7, fig.height=5}
pct_mat <- t(prop.table(tab, margin = 1)) * 100   # subtypes × clusters

barplot(pct_mat,
        col    = subtype_cols[rownames(pct_mat)],  # index by subtype name
        border = NA,
        legend.text = rownames(pct_mat),
        args.legend = list(x = "topright", cex = 0.85),
        ylab = "Tumours (%)", xlab = "Cluster",
        main = "PAM50 subtype composition of each k-means cluster")
```

## Gene expression by cluster

```{r cluster-genes, fig.width=10, fig.height=7}
highlight_genes <- c("ESR1", "ERBB2", "KRT5", "MKI67", "PGR", "FOXA1")
pretty <- c("ESR1 (ER gene)", "ERBB2 (HER2 gene)", "KRT5 (basal)",
            "MKI67 (Ki-67)", "PGR (PR gene)", "FOXA1 (luminal TF)")

par(mfrow = c(2, 3))
for (i in seq_along(highlight_genes)) {
  boxplot(brca[[ highlight_genes[i] ]] ~ scores$cluster,
          col    = cl_cols, border = "grey30",
          main   = pretty[i],
          xlab   = "Cluster", ylab = "log₂ expression")
}
par(mfrow = c(1, 1))
```

> **Question 6.2.** Is any cluster dominated by a single PAM50 subtype?
> Which subtype is most cleanly captured by a single cluster, and which
> subtypes are most likely to be mixed together? Looking at the boxplots:
> which cluster has the highest *ESR1* and *PGR* expression (consistent
> with Luminal A), and which has the highest *KRT5* expression (consistent
> with Basal-like)?

> **Question 6.3.** The clustering algorithm had no access to the PAM50
> subtype labels — it worked only from 50 gene expression values.
> Yet it partially recovers the clinical subtypes. What does this
> demonstrate about the information content of these 50 genes? Does it
> validate the PAM50 test as a robust molecular classification tool?

> **Question 6.4.** Luminal A and Luminal B tumours are both ER-positive,
> yet they have different prognoses and treatment recommendations. If the
> unsupervised clustering tends to group them together, what does this
> suggest about how similar they are at the gene-expression level? What
> additional information — clinical or genomic — might be needed to
> reliably separate them?

---

# Part G — Wrap-up

```{r session}
sessionInfo()
```

> **Question 7.1 (wrap-up).** In four or five sentences, summarise the
> key findings: How well do the first two principal components separate
> the five PAM50 subtypes? What biological programmes do PC1 and PC2
> represent, and which specific genes drive those axes? Does the spatial
> arrangement in PCA space reflect the known clinical hierarchy — with
> Luminal A (best prognosis) and Basal-like (worst prognosis) at opposite
> poles? And what does the unsupervised clustering analysis tell us about
> whether breast cancer gene expression naturally organises into the same
> groups that pathologists and oncologists use clinically?
