---
title: "TP5 — Principal Component Analysis (PCA)"
subtitle: "Doctors as Data Scientists · Dimensionality Reduction and Patient Profiling"
author: "Massimo Andreatta"
date: "`r Sys.Date()`"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file, encoding=encoding,
  output_file=file.path(dirname(input_file), '..', out_dir, 'TP5_PCA_Heart_Failure.html'))})
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, fig.align = "center")
```

> **How to use this notebook.** You do **not** need any programming experience.
> Read each short explanation, then press the green *play* button (▶) at the
> top-right of each grey *code chunk* to run it. Look at the output, and answer
> the **Questions** in the green boxes (write your answers in plain text under
> each one). The goal is to *understand the data and the method*, not to
> memorise code.

# Introduction to Principal Component Analysis

## The problem: too many variables, too few dimensions

In medicine we routinely measure many things about each patient — blood tests,
vital signs, imaging findings. When there are six, ten, or fifty measurements
per patient, it becomes impossible to *see* the data: you cannot draw a
50-dimensional scatter plot. Yet a lot of information in those measurements is
**redundant** — for example, two liver enzymes that always rise together carry
roughly the same information.

**Principal Component Analysis (PCA)** exploits this redundancy to compress
many variables into a small number of new variables — the **principal
components** — that still capture most of the interesting variation. Think of
it as finding the best angle from which to photograph a 3-D sculpture: you
lose one dimension, but if you choose the right angle you preserve the
important structure.

## What PCA actually does — step by step

Imagine you have measured two things about each patient: ejection fraction and
serum creatinine. You could plot every patient as a point in a 2-D scatter
plot. PCA asks: *what single axis through this cloud of points captures the
most spread?* That axis is **PC1**. A second axis, perpendicular to the first,
captures the next most spread — that is **PC2**. Together they form a new
coordinate system that is rotated relative to the original axes.

```{r pca-intuition, fig.width=7, fig.height=5, echo=FALSE}
set.seed(7)
n  <- 120
z1 <- rnorm(n); z2 <- rnorm(n, sd = 0.35)
# Rotate by ~40 degrees to create correlated variables
theta <- -40 * pi / 180
x <- cos(theta) * z1 - sin(theta) * z2
y <- sin(theta) * z1 + cos(theta) * z2

pca_toy <- prcomp(cbind(x, y), scale. = FALSE)
v1 <- pca_toy$rotation[, 1]
v2 <- pca_toy$rotation[, 2]
sc <- 1.5

plot(x, y, pch = 19, col = adjustcolor("steelblue", alpha.f = 0.5), cex = 0.9,
     xlab = "Measurement 1  (e.g. ejection fraction)",
     ylab = "Measurement 2  (e.g. serum creatinine)",
     main = "PCA: finding new axes through the data cloud",
     asp = 1)
arrows(0, 0, sc * v1[1], sc * v1[2],
       col = "red", lwd = 4, length = 0.15)
arrows(0, 0, sc * v2[1], sc * v2[2],
       col = "orange", lwd = 4, length = 0.15)
text(sc * v1[1] * 1.18, sc * v1[2] * 1.18, "PC1", col = "red",    cex = 1.2, font = 2)
text(sc * v2[1] * 1.45, sc * v2[2] * 1.45, "PC2", col = "orange", cex = 1.2, font = 2)
```

The red arrow (PC1) points in the direction of greatest spread — patients are
most spread out along this axis. The orange arrow (PC2) is perpendicular to
PC1 and captures the remaining spread.

In this toy example we still have two dimensions, so nothing is gained yet.
The real benefit appears when we have **many** variables.
PCA finds the directions of greatest variance in the N-dimensional space,
and we can keep only the first two or three PCs — which together capture the
majority of the information — and discard the rest. We compress many dimensions
into two while losing only a small fraction of information, making the data
possible to visualise and interpret.

## Key vocabulary

| Term | Plain-language meaning |
|---|---|
| **Principal component (PC)** | A new axis through the data; a weighted combination of the original variables |
| **Score** | A patient's coordinate on a PC — where they sit along that axis |
| **Loading** | How much each original variable contributes to a PC (its "weight" on that axis) |
| **Variance explained** | The % of total spread captured by a PC — larger means more informative |
| **Scree plot** | A bar chart of variance explained by each PC, used to decide how many to keep |
| **Biplot** | A combined plot showing both patient scores and variable loadings |

## What PCA can — and cannot — do

PCA is an **unsupervised** method: it looks only at the measurements, with no
knowledge of clinical outcomes. It will find the directions of greatest
variation, which may or may not align with what we care about clinically. When
the main sources of biological variation *do* align with disease severity or
prognosis, PCA makes that structure visible. When they do not (for instance,
if the dominant variation is technical noise), PCA can mislead. Always
interpret the components biologically before drawing conclusions.

---

# The dataset: heart failure clinical records

Heart failure occurs when the heart cannot pump enough blood to
meet the needs of the body. When the heart cannot pump properly, the body fails 
to get sufficient oxygen and nutrients.

This dataset contains clinical measurements from
**299 heart failure patients** admitted to a hospital in Faisalabad, Pakistan
(Ahmad T *et al.* 2017, *PLOS ONE*). Each row is one patient; each column is
a clinical measurement or risk factor recorded at admission.

| Variable | Description | Units |
|---|---|---|
| `age` | Age of the patient | years |
| `anaemia` | Decrease of red blood cells (0 = no, 1 = yes) | binary |
| `creatinine_phosphokinase` | CPK enzyme level — marker of muscle damage | mcg/L |
| `diabetes` | Whether the patient has diabetes (0/1) | binary |
| `ejection_fraction` | % of blood pumped out per heartbeat | % |
| `high_blood_pressure` | Hypertension (0/1) | binary |
| `platelets` | Platelet count | kiloplatelets/mL |
| `serum_creatinine` | Serum creatinine — marker of kidney function | mg/dL |
| `serum_sodium` | Serum sodium level | mEq/L |
| `sex` | Sex (0 = female, 1 = male) | binary |
| `smoking` | Smoking status (0/1) | binary |
| `time` | Follow-up period until death or end of study | days |
| `DEATH_EVENT` | Died during follow-up (0 = no, 1 = yes) | binary |

## Load the data

```{r load}
hf <- read.csv("../data/CVD/heart_failure_clinical_records_dataset.csv")
head(hf)
```

---

# Part A — Meet the data

## Structure and variable types

```{r structure}
dim(hf)
str(hf)
```

Some variables are **continuous measurements** (age, ejection fraction,
creatinine…); others are **binary flags** (diabetes, smoking, death event).
PCA works on continuous numbers, so we separate them up front.

```{r split-vars}
# Continuous clinical measurements
cont_vars <- c("age", "creatinine_phosphokinase", "ejection_fraction",
               "platelets", "serum_creatinine", "serum_sodium")

# Binary variables — used later to *interpret* PCA, not as PCA input
bin_vars <- c("anaemia", "diabetes", "high_blood_pressure",
              "sex", "smoking", "DEATH_EVENT")
```

We also **exclude `time`** (follow-up length in days). Its value is partly
determined by *when* a patient died — feeding it into PCA and then colouring
by death outcome would be circular reasoning.

```{r summary}
summary(hf[, cont_vars])
```

> **Question 1.1.** How many patients are in the dataset, and how many
> continuous variables will we use for PCA? What does `ejection_fraction`
> measure, and why might a value below 40 % be clinically worrying?

> **Question 1.2.** From the `summary()` output, which variable has the
> largest numerical range? Does a large range mean that variable is more
> *biologically important*, or is it simply a consequence of the units in
> which it is measured?

## Spotting skewed distributions and log-transforming

Right-skewed variables (a long right tail of very high values) can distort
PCA — a handful of extreme patients would drag the axes. Let us inspect the
two enzyme measurements.

```{r skew-check, fig.width=10, fig.height=3.5}
par(mfrow = c(1, 4))
hist(hf$creatinine_phosphokinase, breaks = 40, col = "steelblue",
     main = "CPK (original)", xlab = "mcg/L")
hist(log(hf$creatinine_phosphokinase), breaks = 40, col = "darkorange",
     main = "log(CPK)", xlab = "log mcg/L")
hist(hf$serum_creatinine, breaks = 40, col = "steelblue",
     main = "Serum creatinine (original)", xlab = "mg/dL")
hist(log(hf$serum_creatinine), breaks = 40, col = "darkorange",
     main = "log(Serum creatinine)", xlab = "log mg/dL")
par(mfrow = c(1, 1))
```

The log transformation makes both distributions far more symmetric. We work
with the log-transformed versions from here on.

```{r log-transform}
hf$log_cpk       <- log(hf$creatinine_phosphokinase)
hf$log_creatinine <- log(hf$serum_creatinine)

cont_clean <- c("age", "log_cpk", "ejection_fraction",
                "platelets", "log_creatinine", "serum_sodium")
```

> **Question 1.3.** Why is it a good idea to log-transform heavily right-skewed
> variables *before* PCA? Hint: think about what happens to the Euclidean
> distance between patients when one variable spans a range of 1–7000 while
> another spans 110–150.

---

# Part B — Why standardise before PCA?

PCA finds the directions of **maximum variance**. If one variable has much
larger numbers than another (e.g. platelet count in thousands vs. serum sodium
in mEq/L), it will dominate the result simply because of its scale — not
because it is more informative.

```{r raw-sd}
round(apply(hf[, cont_clean], 2, sd, na.rm = TRUE), 3)
```

```{r scale}
X <- scale(hf[, cont_clean])    # standardise to mean 0, sd 1

round(apply(X, 2, mean), 3)     # confirm: all ≈ 0
round(apply(X, 2, sd),   3)     # confirm: all = 1
```

After standardisation every variable contributes on equal footing: we compare
*patterns of variation*, not raw magnitudes.

> **Question 2.1.** Before scaling, which variable has the largest standard
> deviation? If we ran PCA on the raw numbers, which variable would likely
> dominate PC1 — and would that reflect biology or just units?

> **Question 2.2.** After standardisation, a patient has a value of +2 for
> `ejection_fraction`. What does that mean in plain English — how does this
> patient compare to the rest of the cohort?

---

# Part C — Running PCA

## Computing the principal components

`prcomp()` rotates the data so that the first axis (**PC1**) captures as much
variance as possible, the second (**PC2**) captures as much of the *remaining*
variance as possible, and so on. The total number of PCs equals the number of
input variables (here, six).

```{r pca}
pca <- prcomp(X, scale. = FALSE)   # already scaled above
summary(pca)
```

The row **Proportion of Variance** tells us how much of the total information
each PC captures.

## Scree plot — how many PCs matter?

```{r scree, fig.width=6, fig.height=4}
var_exp <- pca$sdev^2 / sum(pca$sdev^2) * 100

barplot(var_exp,
        names.arg = paste0("PC", seq_along(var_exp)),
        col = "steelblue", border = NA,
        ylab = "Variance explained (%)",
        main = "Scree plot")
```

We can also plot the **cumulative** variance to decide how many PCs to retain.

```{r cumvar, fig.width=6, fig.height=4}
plot(cumsum(var_exp), type = "b", pch = 19, col = "steelblue",
     xlab = "Number of PCs", ylab = "Cumulative variance explained (%)",
     main = "Cumulative variance explained", ylim = c(0, 100))
abline(h = 80, lty = 2, col = "grey60")
```

> **Question 3.1.** How much variance (%) does PC1 alone explain? How many
> PCs would you need to retain to capture at least 80 % of the total variance
> in the dataset?

> **Question 3.2.** The scree plot shows diminishing returns — each additional
> PC explains less. Where does the curve flatten noticeably (the "elbow")?
> What does this suggest about how many *independent* sources of variation
> exist in these six measurements?

---

# Part D — Patients in PCA space

## The score plot

Each patient becomes a **single point** in PCA space. Points close together
are clinically similar across all six measurements; points far apart differ
substantially. Crucially, the PCA was computed without any knowledge of who
died — we now colour the points by outcome to see whether the structure
aligns with survival.

```{r score-setup}
scores <- as.data.frame(pca$x)
scores$death <- factor(hf$DEATH_EVENT, levels = c(0, 1),
                       labels = c("Survived", "Died"))
```

```{r score-plot, fig.width=6, fig.height=5}
col_death <- ifelse(scores$death == "Died", "tomato", "steelblue")

plot(scores$PC1, scores$PC2,
     col = col_death, pch = 19, cex = 0.8,
     xlab = "PC1", ylab = "PC2",
     main = "PCA score plot — coloured by outcome")
legend("topright", legend = c("Survived", "Died"),
       col = c("steelblue", "tomato"), pch = 19, cex = 0.9)
```

> **Question 4.1.** Do patients who died tend to cluster in a specific region
> of the PC1–PC2 plane? Is the separation between survivors and non-survivors
> perfect, or is there considerable overlap? What does substantial overlap tell
> us about the complexity of predicting heart failure outcomes from these
> measurements alone?

## Mapping individual clinical variables onto the PCA

We can colour the same point cloud by any continuous measurement to reveal
which variable "explains" the position of patients along each PC axis.

```{r color-helper}
# Helper: map numeric values to a red-white-blue gradient
map_color <- function(x, n = 100) {
  pal <- colorRampPalette(c("red", "white", "blue"))(n)
  idx <- cut(x, breaks = n, labels = FALSE, include.lowest = TRUE)
  pal[idx]
}
```

```{r pca-clin, fig.width=10, fig.height=4.5}
par(mfrow = c(1, 3))

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.7,
     col = map_color(hf$ejection_fraction),
     main = "Ejection fraction (%)", xlab = "PC1", ylab = "PC2")

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.7,
     col = map_color(hf$log_creatinine),
     main = "log(Serum creatinine)", xlab = "PC1", ylab = "PC2")

plot(scores$PC1, scores$PC2, pch = 19, cex = 0.7,
     col = map_color(hf$age),
     main = "Age (years)", xlab = "PC1", ylab = "PC2")

par(mfrow = c(1, 1))
```

*In all three plots: red = low values, blue = high values.*

> **Question 4.2.** Looking at the plots coloured by ejection fraction and by
> serum creatinine: do patients with low ejection fraction (red) sit on the
> same side of PC1 as patients with *high* serum creatinine (blue), or the
> opposite side? Recall that a failing heart pumps less blood to the kidneys,
> which can cause creatinine to rise. Does the spatial arrangement make
> clinical sense?

> **Question 4.3.** Does age form a clear gradient along PC1 or PC2? Is it
> surprising that age is not the *dominant* axis of variation in a cohort
> where every patient already has advanced heart failure?

---

# Part E — What drives each PC? Loadings

The **loadings** (stored in `pca$rotation`) tell us how much each original
variable contributes to each PC. A large positive loading means patients with
high values of that variable sit at the *positive* end of the PC; a large
negative loading places them at the *negative* end.

```{r loadings}
round(pca$rotation[, 1:4], 3)
```

```{r loading-bars, fig.width=9, fig.height=4}
par(mfrow = c(1, 2))

barplot(sort(pca$rotation[, 1]),
        horiz = TRUE, las = 1, col = "steelblue", border = NA,
        main = "PC1 loadings", xlab = "Loading value")
abline(v = 0, col = "grey40")

barplot(sort(pca$rotation[, 2]),
        horiz = TRUE, las = 1, col = "darkorange", border = NA,
        main = "PC2 loadings", xlab = "Loading value")
abline(v = 0, col = "grey40")

par(mfrow = c(1, 1))
```

## Biplot — patients and variables together

The **biplot** overlays the score plot with arrows showing each variable's
direction and strength. Arrows pointing the same way indicate variables that
increase together; opposite arrows indicate variables that move in opposite
directions.

```{r biplot, fig.width=7, fig.height=6}
biplot(pca, scale = 0, cex = c(0.45, 0.9),
       col = c("grey60", "tomato"),
       main = "PCA biplot — patients and variable loadings")
```

> **Question 5.1.** Looking at the PC1 loading barplot, which variables have
> the strongest positive and negative loadings? Based on these, propose a
> biological interpretation for PC1: is it primarily an axis of *kidney
> function*, *cardiac function*, or a contrast between the two?

> **Question 5.2.** In the biplot, do the arrows for `ejection_fraction` and
> `log_creatinine` point in roughly *opposite* directions? A failing heart
> pumps less blood to the kidneys, causing creatinine to rise — so patients
> with low ejection fraction often have high serum creatinine. Does the
> geometry of the biplot reflect this clinical relationship?

> **Question 5.3.** Which variables load most strongly on PC2? What might
> PC2 represent? (Hint: look at which variables are *not* well captured by
> PC1 and therefore have most of their signal in PC2.)

---

# Part F — Unsupervised clustering of patients

So far we have only used `DEATH_EVENT` as a colour — the PCA itself was blind
to outcomes. Now we ask: can we find groups of *similar* patients
**without knowing who died**, using only their lab measurements?

**K-means clustering** assigns each patient to one of *k* groups so that
patients in the same group are as similar as possible. We apply it to the
first four PCs (which already capture most of the variance).

## Choosing the number of clusters

The *within-cluster sum of squares* (WSS) measures how tight the clusters
are. It always drops as *k* increases, so we look for the "elbow" — the point
where adding more clusters gives diminishing returns.

```{r elbow, fig.width=6, fig.height=4}
set.seed(42)
wss <- sapply(1:8, function(k) {
  kmeans(pca$x[, 1:4], centers = k, nstart = 25)$tot.withinss
})

plot(1:8, wss, type = "b", pch = 19, col = "steelblue",
     xlab = "Number of clusters (k)",
     ylab = "Total within-cluster sum of squares",
     main = "Elbow plot for k-means")
```

> **Question 6.1.** The WSS always decreases as *k* grows — so we cannot
> just maximise it. Where does the elbow appear in the curve? What value of
> *k* offers a good balance between simplicity (few clusters) and fit (tight
> clusters)?

## Clustering with k = 2

```{r kmeans}
set.seed(42)
km <- kmeans(pca$x[, 1:4], centers = 2, nstart = 25)
scores$cluster <- as.factor(km$cluster)
```

```{r cluster-plot, fig.width=6, fig.height=5}
col_cl <- c("1" = "#1f78b4", "2" = "#e66101")

plot(scores$PC1, scores$PC2,
     col = col_cl[scores$cluster], pch = 19, cex = 0.8,
     xlab = "PC1", ylab = "PC2",
     main = "K-means clusters (k = 2) in PCA space")
legend("topright", legend = c("Cluster 1", "Cluster 2"),
       col = col_cl, pch = 19, cex = 0.9)
```

## Are the unsupervised clusters enriched in death events?

```{r cluster-vs-death}
tab <- table(Cluster = scores$cluster, Outcome = scores$death)
tab
round(prop.table(tab, margin = 1) * 100, 1)
```

```{r cluster-bar, fig.width=5, fig.height=4.5}
barplot(t(prop.table(tab, margin = 1)) * 100,
        col = c("steelblue", "tomato"),
        legend.text = c("Survived", "Died"),
        args.legend = list(x = "topright", cex = 0.9),
        ylab = "Patients (%)", xlab = "Cluster",
        main = "Death rate by unsupervised cluster",
        border = NA)
```

## Clinical profile of each cluster

```{r cluster-profile, fig.width=10, fig.height=7}
# Friendly labels for the plot
pretty_names <- c("Age (years)", "log(CPK)", "Ejection fraction (%)",
                  "Platelets (kplt/mL)", "log(Serum creatinine)", "Serum sodium (mEq/L)")

par(mfrow = c(2, 3))
for (i in seq_along(cont_clean)) {
  boxplot(hf[[ cont_clean[i] ]] ~ scores$cluster,
          col = c("#1f78b4", "#e66101"), border = "grey30",
          main = pretty_names[i], xlab = "Cluster", ylab = "")
}
par(mfrow = c(1, 1))
```

> **Question 6.2.** Is one cluster substantially enriched for patients who
> died? Looking at the clinical profile boxplots, which cluster has *lower*
> ejection fraction and *higher* serum creatinine — and is that the same
> cluster with more deaths?

> **Question 6.3.** The clusters were defined purely by six lab measurements,
> with no knowledge of who survived. What does it mean if unsupervised
> clusters still align reasonably well with survival? Does this suggest that
> routine blood tests carry clinically relevant prognostic information?

---

# Part G — Wrap-up

```{r session}
sessionInfo()
```

> **Question 7.1 (wrap-up).** In four or five sentences, summarise the key
> findings: What are the main axes of variation among these heart failure
> patients? Which clinical measurements drive those axes, and how do they
> relate to each other biologically? Do patients who died occupy a distinct
> region of PCA space? And what do the unsupervised clusters reveal about the
> structure of the cohort?
