---
title: "TP7 — Unsupervised Clustering"
subtitle: "K-means, hierarchical clustering, and cluster validation - Breast cancer subtypes"
author: "Massimo Andreatta"
date: "`r Sys.Date()`"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file, encoding=encoding,
  output_file=file.path(dirname(input_file), '..', out_dir, 'TP7_Clustering.html'))})
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, fig.align = "center")
```

> **How to use this notebook.** You do **not** need any programming experience.
> Read each short explanation, then press the green *play* button (▶) at the
> top-right of each grey *code chunk* to run it. Look at the output, and answer
> the **Questions** in the green boxes (write your answers in plain text under
> each one). The goal is to *understand the data and the method*, not to
> memorise code.

> **Building on TP4–TP6.** This notebook assumes you have completed TP4–TP6,
> which introduced PCA using three different datasets and culminated in a
> full PCA analysis of a breast cancer gene-expression cohort (TP6). TP6 ended
> with a first taste of **unsupervised clustering** using k-means; here we
> slow down, cover clustering properly, add **hierarchical clustering**, and
> learn how to **validate** a clustering result rather than just eyeballing a
> plot.

---

# From PCA to clustering

## Two different questions

PCA and clustering are both **unsupervised** — neither uses the diagnosis or
subtype labels. But they answer different questions:

- **PCA** asks: *what are the axes along which patients/tumours vary the
  most?* Its output is a set of continuous new coordinates (PC1, PC2, …).
- **Clustering** asks: *which points are similar enough to be grouped
  together?* Its output is a discrete label — cluster 1, cluster 2, … — for
  every point.

The two are often used together, exactly as we will do here: run PCA first
to compress many genes into a handful of informative axes, then cluster the
patients using those axes rather than the original noisy, high-dimensional
data.

## Key vocabulary

| Term | Plain-language meaning |
|---|---|
| **Cluster** | A group of points considered similar to one another |
| **Distance / dissimilarity** | A number quantifying how different two points are — the input to every clustering method |
| **Centroid** | The "average" point of a cluster (used by k-means) |
| **Dendrogram** | A tree diagram showing how points are progressively merged in hierarchical clustering |
| **Linkage** | The rule used to decide how far apart two *clusters* are, based on the points inside them |
| **Within-cluster sum of squares (WSS)** | How spread out the points are within their assigned cluster — lower is tighter |
| **Silhouette width** | A score (from −1 to +1) for how well a point fits its own cluster compared with the next-closest cluster |

## A toy example: how k-means finds groups

Before working with a real dataset, let's build intuition with two simple made-up
measurements and three obviously separated groups of points.

```{r toy-kmeans, fig.width=7, fig.height=5.5, echo=FALSE}
set.seed(11)
n_per <- 40
blob1 <- cbind(rnorm(n_per, mean = 0,   sd = 0.7), rnorm(n_per, mean = 0,   sd = 0.7))
blob2 <- cbind(rnorm(n_per, mean = 5,   sd = 0.7), rnorm(n_per, mean = 0,   sd = 0.7))
blob3 <- cbind(rnorm(n_per, mean = 2.5, sd = 0.7), rnorm(n_per, mean = 4.3, sd = 0.7))
toy <- rbind(blob1, blob2, blob3)

km_toy   <- kmeans(toy, centers = 3, nstart = 25)
toy_cols <- c("#e41a1c", "#377eb8", "#4daf4a")

plot(toy, pch = 19, cex = 1.2,
     col  = toy_cols[km_toy$cluster],
     xlab = "Measurement 1", ylab = "Measurement 2",
     main = "K-means finds groups of nearby points")
points(km_toy$centers, pch = 4, cex = 2.5, lwd = 3, col = "black")
legend("topright", legend = c("Cluster 1", "Cluster 2", "Cluster 3", "Centroid"),
       col = c(toy_cols, "black"), pch = c(19, 19, 19, 4), pt.cex = c(1, 1, 1, 1.5))
```

K-means starts from random guesses for the cluster centres, assigns every
point to its nearest centre, recomputes the centres as the average of their
assigned points, and repeats until nothing changes. The black crosses are the
final **centroids** — the "centre of gravity" of each cluster.

> **Question 0.1.** In the toy plot above, the three groups are very
> obviously separated. Do you think k-means would still find sensible
> clusters if the three blobs overlapped substantially? What would happen to
> points that fall in the region where two blobs overlap?

> **Question 0.2.** K-means requires you to specify the number of clusters,
> *k*, in advance. In the toy example we told it `centers = 3`. What do you
> think would happen if we asked k-means for `centers = 2` on this same
> data? What about `centers = 6`?

---

# The dataset: breast cancer gene expression (recap)

We continue with the same **TCGA-BRCA** gene-expression dataset used in TP6:
522 tumours, 50 PAM50 genes, and a known PAM50 subtype label (LumA, LumB,
Her2, Basal, NormalLike) for each tumour. As in TP6, we standardise the genes
and run PCA — the code below reproduces that pipeline without repeating the
explanations (see TP6 for the full walk-through).

```{r load}
brca <- read.csv("../data/brca_pam50.csv", row.names = 1)
gene_cols <- colnames(brca)[7:56]

subtype_cols <- c(
  LumA       = "#1f77b4",
  LumB       = "#aec7e8",
  Her2       = "#ff7f0e",
  Basal      = "#d62728",
  NormalLike = "#7f7f7f"
)

X   <- scale(brca[, gene_cols])
pca <- prcomp(X, scale. = FALSE)
var_exp <- pca$sdev^2 / sum(pca$sdev^2) * 100

scores <- as.data.frame(pca$x)
scores$subtype <- factor(brca$pam50,
                         levels = c("LumA", "LumB", "Her2", "Basal", "NormalLike"))
col_pts <- subtype_cols[as.character(scores$subtype)]
```

```{r score-plot-recap, fig.width=7, fig.height=5.5}
plot(scores$PC1, scores$PC2,
     col  = col_pts, pch = 19, cex = 0.7,
     xlab = paste0("PC1 (", round(var_exp[1], 1), " %)"),
     ylab = paste0("PC2 (", round(var_exp[2], 1), " %)"),
     main = "PCA plot — coloured by PAM50 subtype (recap from TP6)")
legend("topright", legend = names(subtype_cols), col = subtype_cols,
       pch = 19, cex = 0.85, pt.cex = 1.1)
```

From here on, we will **not use the `subtype` colours to build the
clusters** — clustering only sees the PCA coordinates. We use the known
labels afterwards, purely to check whether the clusters we find agree with
the known biology.

We cluster tumours using the first five PCs (which together capture most of
the informative variance — see the scree plot in TP6) rather than the raw
50 genes: this reduces noise and speeds up computation.

```{r npc}
npc <- 5   # number of PCs used as clustering input
```

---

# 1 — K-means clustering

Can we **rediscover** the PAM50 subtypes **without** using the labels — only
from the 50 gene expression values (summarised by their first 5 PCs)?

**K-means clustering** assigns each tumour to one of *k* groups so that
tumours in the same group are as similar as possible to their group's
centroid — the same idea as the toy example above, now applied to real data.

## Choosing the number of clusters (elbow plot)

```{r elbow, fig.width=6, fig.height=4}
set.seed(42)

wss <- sapply(1:10, function(k) {
  kmeans(pca$x[, 1:npc], centers = k, nstart = 25)$tot.withinss
})

plot(1:10, wss, type = "b", pch = 19, col = "steelblue",
     xlab = "Number of clusters (k)",
     ylab = "Total within-cluster sum of squares",
     main = "Elbow plot for k-means")
```

> **Question 1.1.** There are five known PAM50 subtypes (though Normal-like
> has only 8 samples). Does the elbow plot suggest *k* = 5 as a clear
> inflection point, or does the curve flatten earlier? What might this tell
> us about whether all five subtypes form truly distinct clusters in gene
> expression space?

## Clustering with k = 4

We choose *k* = 4 to match the four major clinical subtypes (LumA, LumB,
Her2, Basal), setting aside the rare Normal-like category — and to allow a
direct comparison with the hierarchical clustering below.

```{r kmeans}
set.seed(42)
cl_cols <- c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3")

km <- kmeans(pca$x[, 1:npc], centers = 4, nstart = 25)
scores$cluster_km <- as.factor(km$cluster)
```

```{r cluster-plot, fig.width=7, fig.height=5.5}
plot(scores$PC1, scores$PC2,
     col = cl_cols[scores$cluster_km], pch = 19, cex = 0.7,
     xlab = paste0("PC1 (", round(var_exp[1], 1), " %)"),
     ylab = paste0("PC2 (", round(var_exp[2], 1), " %)"),
     main = "K-means clusters (k = 4) in PCA space")
legend("topright", legend = paste("Cluster", 1:4),
       col = cl_cols, pch = 19, cex = 0.85)
```

## How well do the k-means clusters match PAM50 subtypes?

```{r confusion}
tab_km <- table(Cluster = scores$cluster_km, PAM50 = scores$subtype)
tab_km
round(prop.table(tab_km, margin = 1) * 100, 1)
```

```{r cluster-bar, fig.width=7, fig.height=5}
pct_mat <- t(prop.table(tab_km, margin = 1)) * 100   # subtypes × clusters

barplot(pct_mat,
        col    = subtype_cols[rownames(pct_mat)],
        border = NA,
        legend.text = rownames(pct_mat),
        args.legend = list(x = "topright", cex = 0.85),
        ylab = "Tumours (%)", xlab = "Cluster",
        main = "PAM50 subtype composition of each k-means cluster")
```

## Gene expression by cluster

```{r cluster-genes, fig.width=10, fig.height=7}
highlight_genes <- c("ESR1", "ERBB2", "KRT5", "MKI67", "PGR", "FOXA1")
pretty <- c("ESR1 (ER gene)", "ERBB2 (HER2 gene)", "KRT5 (basal)",
            "MKI67 (Ki-67)", "PGR (PR gene)", "FOXA1 (luminal TF)")

par(mfrow = c(2, 3))
for (i in seq_along(highlight_genes)) {
  boxplot(brca[[ highlight_genes[i] ]] ~ scores$cluster_km,
          col    = cl_cols, border = "grey30",
          main   = pretty[i],
          xlab   = "Cluster", ylab = "log₂ expression")
}
par(mfrow = c(1, 1))
```

> **Question 1.2.** Is any cluster dominated by a single PAM50 subtype?
> Which subtype is most cleanly captured by a single cluster, and which
> subtypes are most likely to be mixed together? Looking at the boxplots:
> which cluster has the highest *ESR1* and *PGR* expression (consistent
> with Luminal A), and which has the highest *KRT5* expression (consistent
> with Basal-like)?

> **Question 1.3.** The clustering algorithm had no access to the PAM50
> subtype labels — it worked only from gene expression values. Yet it
> partially recovers the clinical subtypes. What does this demonstrate about
> the information content of these 50 genes?

> **Question 1.4.** Luminal A and Luminal B tumours are both ER-positive,
> yet they have different prognoses and treatment recommendations. If the
> unsupervised clustering tends to group them together, what does this
> suggest about how similar they are at the gene-expression level?

---

# 2 — Hierarchical clustering

## How it works

Hierarchical clustering does not require choosing *k* in advance. It starts
with every point in its own cluster, then repeatedly **merges the two
closest clusters** until only one giant cluster remains. The record of every
merge is stored in a **dendrogram** — a tree you can "cut" at any height to
obtain any number of clusters.

Two ingredients decide the result:

- **Distance metric** — how do we measure how far apart two individual
  tumours are? We use **Euclidean distance** on the first five PCs.
- **Linkage** — how do we measure the distance between two *clusters* (each
  containing several tumours), not just two points? We use **Ward's
  method**, which merges clusters in the way that increases within-cluster
  variance the least — it tends to produce compact, similarly sized
  clusters.

```{r hclust}
d  <- dist(pca$x[, 1:npc])
hc <- hclust(d, method = "ward.D2")
```

## The dendrogram

```{r dendrogram, fig.width=9, fig.height=5}
plot(hc, labels = FALSE, hang = -1,
     main = "Hierarchical clustering dendrogram (522 tumours)",
     xlab = "Tumours", sub = "")
rect.hclust(hc, k = 4, border = cl_cols)
```

Each leaf at the bottom is one tumour (too many to label individually); the
height of each merge shows how dissimilar the two merging groups were. The
coloured boxes show where the tree is cut to produce **4 clusters** — the
same number we used for k-means above, so the two methods can be compared
fairly.

> **Question 2.1.** Looking at the dendrogram, do the four coloured boxes
> split off at similar heights, or does one box separate from the rest much
> earlier (at a much greater height) than the others? What would that tell
> you about how distinct that group is from the rest?

> **Question 2.2.** If you cut the dendrogram at a much lower height (close
> to 0), would you get more or fewer clusters than 4? What about cutting it
> at the very top, just below the root?

## Hierarchical clusters in PCA space

```{r hclust-cut}
scores$cluster_hc <- as.factor(cutree(hc, k = 4))
```

```{r hclust-plot, fig.width=7, fig.height=5.5}
plot(scores$PC1, scores$PC2,
     col  = cl_cols[scores$cluster_hc], pch = 19, cex = 0.7,
     xlab = paste0("PC1 (", round(var_exp[1], 1), " %)"),
     ylab = paste0("PC2 (", round(var_exp[2], 1), " %)"),
     main = "Hierarchical clusters (k = 4) in PCA space")
legend("topright", legend = paste("Cluster", 1:4), col = cl_cols,
       pch = 19, cex = 0.85)
```

## Do hierarchical clusters match the known PAM50 subtypes?

```{r hclust-confusion}
tab_hc <- table(Cluster = scores$cluster_hc, PAM50 = scores$subtype)
tab_hc
round(prop.table(tab_hc, margin = 1) * 100, 1)
```

> **Question 2.3.** Compare the hierarchical-clustering PCA plot with the
> PAM50-coloured PCA plot from the recap section. Does one hierarchical
> cluster correspond closely to a single PAM50 subtype? Which subtype is
> "easiest" for hierarchical clustering to isolate?

> **Question 2.4.** Ward's linkage tends to produce clusters of similar
> size. Looking at the cluster sizes in the confusion table, does this
> assumption seem appropriate for this dataset — are the true PAM50 subtypes
> (see the bar chart in TP6, section 1) roughly equal in size, or very
> unequal? Could a "prefers equal-sized clusters" method be a poor fit here?

---

# 3 — Comparing the two clustering methods

K-means and hierarchical clustering start from very different logic (an
iteratively updated set of centroids vs. a merging tree), but we ran both
with the same *k* = 4 on the same 5-dimensional PCA input. How similar are their
results?

```{r compare-methods}
tab_compare <- table(`K-means` = scores$cluster_km, `Hierarchical` = scores$cluster_hc)
tab_compare
```

```{r compare-plot, fig.width=12, fig.height=5.5}
par(mfrow = c(1, 2))

plot(scores$PC1, scores$PC2,
     col = cl_cols[scores$cluster_km], pch = 19, cex = 0.6,
     xlab = "PC1", ylab = "PC2", main = "K-means (k = 4)")

plot(scores$PC1, scores$PC2,
     col = cl_cols[scores$cluster_hc], pch = 19, cex = 0.6,
     xlab = "PC1", ylab = "PC2", main = "Hierarchical clustering (k = 4)")

par(mfrow = c(1, 1))
```

> **Question 3.1.** Look at the cross-tabulation above. Is there roughly a
> **one-to-one correspondence** between k-means clusters and hierarchical
> clusters (each k-means cluster maps almost entirely onto one hierarchical
> cluster), or is the agreement weaker than that? Note that cluster
> *numbers* are arbitrary and may not match between methods — focus on
> whether entire *groups* of tumours agree, not on matching numeric labels.

> **Question 3.2.** Where the two methods *disagree*, which tumours do you
> think are involved — ones near the centre of a cluster, or ones near the
> boundary between two clusters? Why would boundary tumours be the ones most
> likely to be assigned differently by two different algorithms?

---

# 4 — How many clusters? Silhouette validation

The elbow plot in section 1 is useful but subjective — there is no single
"correct" answer for where the elbow is. The **silhouette width** gives a
more objective, per-point score.

## What silhouette width measures

For each point, the silhouette width compares:

- **a** = the average distance to other points in its *own* cluster (lower
  is better — the point fits well where it is)
- **b** = the average distance to points in the *nearest other* cluster
  (higher is better — the point is far from its nearest alternative)

The silhouette width is `(b - a) / max(a, b)`, ranging from **−1** (probably
in the wrong cluster) to **+1** (very well matched to its cluster). Averaging
this score across all points gives a single number to compare different
values of *k*.

```{r sil-choose-k, fig.width=6, fig.height=4}
library(cluster)

sil_width <- sapply(2:10, function(k) {
  cl <- kmeans(pca$x[, 1:npc], centers = k, nstart = 25)$cluster
  mean(silhouette(cl, d)[, "sil_width"])
})

plot(2:10, sil_width, type = "b", pch = 19, col = "darkorange",
     xlab = "Number of clusters (k)",
     ylab = "Average silhouette width",
     main = "Silhouette validation for k-means")
```

> **Question 4.1.** Which value of *k* has the highest average silhouette
> width? Is it the same value suggested by the elbow plot, or by the known
> number of PAM50 subtypes (5, or 4 if Normal-like is excluded)? If they
> disagree, which one would you trust more, and why?

## Silhouette plot for k = 4

We can also look at silhouette width **per tumour**, grouped by cluster, to
see whether some clusters are much better defined than others.

```{r sil-plot, fig.width=7, fig.height=6}
sil_km <- silhouette(km$cluster, d)

par(mar = c(4, 4, 4, 8))
plot(sil_km, col = cl_cols, border = NA,
     main = "Silhouette plot — k-means, k = 4", cex.names = 0.7)
```

Each horizontal bar is one tumour; bars are grouped and sorted by cluster.
Tumours with a **negative** silhouette width are, on average, closer to a
*different* cluster than the one they were assigned to — a sign of a
poorly-defined boundary or a genuinely intermediate tumour.

> **Question 4.2.** Which cluster has the highest average silhouette width
> (shown in the per-cluster label), and which has the lowest? Cross-reference
> with the PAM50 confusion table from section 1 — is the cluster with the
> *highest* silhouette width also the one most dominated by a single PAM50
> subtype?

> **Question 4.3.** Are there any tumours with a negative silhouette width?
> Based on what you learned about Luminal A and Luminal B being biologically
> similar, would you expect tumours from these two subtypes to be
> over-represented among the poorly-clustered (low or negative silhouette)
> points?

## Internal vs. external validation: checking against known labels

Silhouette width (like the elbow plot) is an **internal** validity measure —
it only looks at the geometry of the clusters themselves (how compact, how
well-separated), with no idea whether the resulting groups mean anything
biologically. Because we happen to know the true PAM50 subtype of every
tumour, we can also compute an **external** validity measure: how well does
each clustering solution agree with the known labels? In most real
unsupervised analyses labels do not exist, so this check is a luxury — but it
is a good way to see what internal measures can and cannot tell you.

The **Adjusted Rand Index (ARI)** compares two groupings of the same points.
It scores **1** for a perfect match, **~0** for agreement no better than
random chance, and negative values for worse-than-chance agreement.

```{r ari-function}
# Adjusted Rand Index: agreement between two clusterings of the same points
# (1 = perfect match, ~0 = no better than random)
adjusted_rand_index <- function(cl1, cl2) {
  tab     <- table(cl1, cl2)
  choose2 <- function(x) x * (x - 1) / 2
  n       <- sum(tab)

  index    <- sum(choose2(tab))
  expected <- sum(choose2(rowSums(tab))) * sum(choose2(colSums(tab))) / choose2(n)
  maxindex <- 0.5 * (sum(choose2(rowSums(tab))) + sum(choose2(colSums(tab))))

  (index - expected) / (maxindex - expected)
}
```

```{r ari-by-k, fig.width=6, fig.height=4}
set.seed(42)
ari_by_k <- sapply(2:10, function(k) {
  cl <- kmeans(pca$x[, 1:npc], centers = k, nstart = 25)$cluster
  adjusted_rand_index(cl, brca$pam50)
})

plot(2:10, ari_by_k, type = "b", pch = 19, col = "seagreen",
     xlab = "Number of clusters (k)",
     ylab = "Adjusted Rand Index vs. true PAM50 label",
     main = "External validation: agreement with known subtypes")
```

> **Question 4.4.** Compare the three curves you have now produced: the
> elbow plot (section 1), the silhouette-vs-*k* plot (above), and this ARI
> plot. Do they point to the same value of *k*, or to different ones? The
> elbow and silhouette curves are **internal** measures — they never see the
> PAM50 labels. ARI is **external** — it directly checks agreement against
> the labels. The silhouette width was highest at *k* = 2; can you think of
> a biological reason why the single largest split in these tumours might be
> a two-way one (hint: think about oestrogen-receptor status)? What does the
> fact that internal measures do not necessarily peak at the "clinically
> correct" number of clusters tell you about the limits of choosing *k* in a
> purely unsupervised way, with no external knowledge at all?

---

# 5 — Wrap-up

```{r session}
sessionInfo()
```

> **Question 5.1 (wrap-up).** In four or five sentences, summarise the
> journey: How do k-means and hierarchical clustering compare on this
> dataset — do they broadly agree? How well does either method recover the
> known PAM50 subtypes, and which subtype is easiest / hardest to isolate
> without labels? The elbow plot, silhouette width, and ARI each suggested a
> different "best" *k* — what does that disagreement teach you about
> internal vs. external cluster validation? Finally, based on everything in
> TP4–TP7, what do you see as the main value of unsupervised methods (PCA +
> clustering) in a clinical/biological research setting where labels are not
> always available?
